### 2019-07-03 paper.tex
first push

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{tikz}
\usetikzlibrary{arrows,shapes,positioning,shadows,trees}
\tikzset{
  basic/.style  = {draw, text width=2cm, drop shadow, font=\sffamily, rectangle},
  root/.style   = {basic, rounded corners=2pt, thin, align=center,
                   fill=green!30},
  level 2/.style = {basic, rounded corners=6pt, thin,align=center, fill=green!60,
                   text width=8em},
  level 3/.style = {basic, thin, align=left, fill=pink!60, text width=6.5em}
}

\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{{img/}}
\usepackage{tabularx}
\newcommand{\tableheadline}[1]{\multicolumn{1}{c}{#1}}
\usepackage{dirtree}
\usepackage{rotating}
\usepackage{caption}
\usepackage{booktabs}
% More space between rows:		
\renewcommand{\arraystretch}{1.1}		
% Slightly more space between columns:		
\setlength{\tabcolsep}{8pt}
\usepackage{minted}
\definecolor{dhscodebg}{rgb}{0.95,0.95,0.95} % listing background

\usepackage{xspace}
\renewcommand\UrlFont{\color{blue}\rmfamily}

\newcommand{\mypddl}{\textsc{myPddl}\xspace}
\newcommand{\mypddlclojure}{\textsc{myPddl-clojure}\xspace}
\newcommand{\mypddlsyntax}{\textsc{myPddl-syntax}\xspace}
\newcommand{\mypddldiagram}{\textsc{myPddl-diagram}\xspace}
\newcommand{\mypddlnew}{\textsc{myPddl-new}\xspace}
\newcommand{\mypddlide}{\textsc{myPddl-ide}\xspace}
\newcommand{\mypddlsnippet}{\textsc{myPddl-snippet}\xspace}
\newcommand{\mypddldistance}{\textsc{myPddl-distance}\xspace}
\newcommand{\mypddlplan}{\textsc{myPddl-plan}\xspace}
\newcommand{\pddlstudio}{\textsc{pddl studio}\xspace}
\newcommand{\itsimple}{\textsc{itSimple}\xspace}
\newcommand{\pddlmode}{\textsc{pddl}-mode\xspace}
\newcommand{\pddl}{\textsc{pddl}\xspace}
\newcommand{\uml}{\textsc{uml}\xspace}
\newcommand{\ide}{\textsc{ide}\xspace}
\newcommand{\sublimetext}{Sublime Text\xspace}

\begin{document}
%
\title{MyPDDL: Tools for efficiently creating PDDL domains and
  problems}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Volker Strobel\inst{1}\orcidID{0000-0003-2974-9827} \and
Alexandra Kirsch\inst{2}\orcidID{0000−0002−5663−1798}}
%
\authorrunning{V. Strobel and A. Kirsch}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{IRIDIA, Universit\'e Libre de Bruxelles, Belgium \and
Independent Scientist\\
\email{vstrobel@ulb.ac.be}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
  The Planning Domain Definition Language (PDDL) is the
  state-of-the-art language for specifying planning problems in
  artificial intelligence research. Writing and maintaining these
  planning problems, however, can be time-consuming and error
  prone. To address this issue, we present myPDDL---a modular toolkit
  for developing and manipulating PDDL domains and problems. To
  evaluate myPDDL, we compare its features to existing knowledge
  engineering tools for PDDL. In a user test, we additionally assess
  two of its modules, namely the syntax highlighting feature and the
  type diagram generator. The users of syntax highlighting detected
  36\,\% more errors than non-users in an erroneous domain file. The
  average time on task for questions on a PDDL type hierarchy was
  reduced by 48\,\% when making the type diagram generator
  available. This implies that myPDDL can support knowledge engineers
  well in the PDDL design and analysis process.

\keywords{PDDL \and Planning}
\end{abstract}
%
%

\section{Introduction}
\label{sec:introduction}

Being a key aspect of artificial intelligence, \emph{planning} is
concerned with devising a sequence of actions to achieve a desired
goal \cite{helmert2008understanding}. It can be both a tool to create
automated systems and a means to support and understand human behavior
\cite{konar1999artificial}. However, the effectiveness of planning
largely depends on the quality of the problem formalization
\cite{shah2013knowledge,keps2014}. To ensure a standardized modeling
format, the Planning Domain Definition Language (\textsc{pddl})
\cite{mcdermott1998pddl} was developed and has become the de facto
standard for the description of planning tasks
\cite{ilghami2005extension}. The discipline that deals with the
integration of world information into a computer system via a human
expert is called knowledge engineering \cite{feigenbaum1983fifth}.
While automated planning could save vast amounts of time if everything
works as intended, creating the planning task specifications is a
complex task that can be error-prone and cumbersome.

The specification of Artificial Intelligence (AI) planning can be
time-consuming and cubersome. While there is a large spectrum of
different AI languages, the de facto standard language is PDDL
(\emph{Planning Domain Definition Language}). However, so far its use
is predominantly confined to academic labs. One of the reasons seems
to be the missing support of knowledge engieering tools so simplify
the task of specifying complex planning domains.

In this chapter, we describe \mypddl
(Figure~\ref{fig:mypddl-overview}), a knowledge engineering toolkit
that supports the AI knowledge engineer in the development of complex
AI planning problems. \mypddl is intended to support knowledge
engineers throughout the entire design cycle of specifying planning
tasks. In the initial stages, it allows for the creation of structured
\textsc{pddl} projects that should encourage a disciplined design
process. With the help of snippets, i.e. code templates, often used
constructs can be inserted in \textsc{pddl} files. A syntax
highlighting feature that speeds up the error-detection can come in
handy in intermediate stages. Understanding the textual representation
of complex type hierarchies in domain files can be confusing, so an
additional tool enables their visualization. \textsc{pddl}'s limited
modeling capabilities were bypassed by developing an interface that
converts \pddl code into Clojure \cite{hickey2008clojure} code and
vice versa.  Within this project, the interface was employed for a
feature that calculates distances between objects specified in a
problem model, but the interface provides numerous other possibilities
and could also be used to further automate the modeling process. All
of the features were integrated into the customizable and extensible
Sublime Text \cite{sublimetext3} editor. Since the main aim in the
development of the toolkit was for it to be easy to use and maintain,
it is evaluated with regard to these criteria. The usability was
assessed by means of a user test with eight subjects that had no prior
experience with artificial intelligence planning. The results indicate
that both error-detection and the understanding of a given domain can
be facilitated by \mypddl.

\begin{figure}[h]
  \centering
  \begin{tikzpicture}[
  level 1/.style={sibling distance=40mm},
  edge from parent/.style={->,draw},
  >=latex]

  % root of the the initial tree, level 1
  \node[] (ide) {-IDE};
\node[below of = ide,root] {myPDDL}
% The first level, as children of the initial tree
  child {node[level 2] (c1) {-new}}
  child {node[level 2] (c2) {-syntax}}
  child {node[level 2] (c3) {-snippet}}
  child {node[level 2] (c4) {-distance}}
  child {node[level 2] (c5) {-diagram}};

% The second level, relatively positioned nodes
\begin{scope}[every node/.style={level 3}]
\node [below of = c1, xshift=15pt] (c11) {Setting shape};
\node [below of = c11] (c12) {Choosing color};
\node [below of = c12] (c13) {Adding shading};

\node [below of = c2, xshift=15pt] (c21) {Using a Matrix};
\node [below of = c21] (c22) {Relatively};
\node [below of = c22] (c23) {Absolutely};
\node [below of = c23] (c24) {Using overlays};

\node [below of = c3, xshift=15pt] (c31) {Default arrows};
\node [below of = c31] (c32) {Arrow library};
\node [below of = c32] (c33) {Resizing tips};
\node [below of = c33] (c34) {Shortening};
\node [below of = c34] (c35) {Bending};
\end{scope}

% lines from each level 1 node to every one of its "children"
\foreach \value in {1,2,3}
  \draw[->] (c1.195) |- (c1\value.west);

\foreach \value in {1,...,4}
  \draw[->] (c2.195) |- (c2\value.west);

\foreach \value in {1,...,5}
  \draw[->] (c3.195) |- (c3\value.west);
\end{tikzpicture}
  \caption{\mypddl is a highly customizable and extensible modular
    system, designed for supporting knowledge engineers in the process
    of writing, analyzing and expanding PDDL files and thereby
    promoting the collaboration between knowledge engineers and the
    use of PDDL in real-world applications. It consists of the parts
    shown in the figure.}
  \label{fig:mypddl-overview}
\end{figure}

The remainder of this chapter is structured as
follows. Section~\ref{sec:related-work} compares other PDDL editors
and \pddl knowledge engineering tools. Section~\ref{sec:myPDDL}
describes the different modules of \mypddl and their design
principles. Section~\ref{sec:valid-eval} compares \mypddl to the other
existing tools using a benchmark
validation. Section~\ref{sec:conclusion} concludes the chapter and
outlines future work.

\section{Related Work}
\label{sec:related-work}

This section introduces knowledge engineering tools that allow editing
\textsc{pddl} files in a textual environment. After introducing the
tools, they are compared and their shortcomings are discussed to set
the stage for \mypddl.

\pddlstudio \cite{plch2012inspect} is an application for creating and
managing \pddl projects, i.e. a collection of \pddl
files. \pddlstudio 's integrated development environment (\ide) was
inspired by Microsoft Visual Studio and imperative programming
paradigms. Its main features are syntax highlighting, error detection,
context sensitive code completion, code folding, project management,
and planner integration. \pddlstudio 's error detection can recognize
both syntactic (missing keywords, parentheses, etc.) and semantic
(wrong type of predicate parameters, misspelled predicates, etc.)
errors.

A major drawback of \pddlstudio is that it is not updated regularly
and only supports \pddl 1.2. Later \pddl versions contain several
additional features such as durative actions, numeric fluents, and
plan metrics \cite{edelkamp2004pddl2}.

\itsimple \cite{vaquero2005itsimple} follows a graphical approach
using Unified Modeling Language (\uml) diagrams. In the process
leading up to \itsimple, \textsc{uml.p} (\uml in a Planning Approach)
was proposed, a \uml variant specifically designed for modeling
planning domains and problems \cite{vaquero2006use}.

\itsimple's modeling workflow is unidirectional as changes in the
\pddl domain do not affect the \uml model and \uml models have to be
modeled manually, meaning that they cannot by generated from
\pddl. However, \cite{tonidandel2006reading} present a translation
process from a \pddl domain specification to an object-oriented
\textsc{uml.p} model as a possible integration for \itsimple. This
translation process makes extensive semantic assumptions for \pddl
descriptions. For example, the first parameter in the
\texttt{:parameters} section of an action is automatically declared as
a subclass of the default class \texttt{Agent}, and the method is
limited to predicates with a maximum arity of two. The current
version of \itsimple does not include the translation process from
\pddl to \uml.

Starting in version 4.0, \itsimple expanded its features to allow the
creation of \pddl projects from scratch (i.e. without the \uml to
\pddl translation process) \cite{vaquero2012itsimple4}. Thus far, the
\pddl editing features are basic. A minimal syntax highlighting
feature recognizes \pddl keywords, variables, and comments. \itsimple
also provides templates for \pddl constructs, such as requirement
specifications, predicates, actions, initial state, and goal
definitions.

Both \pddlstudio and \itsimple do not build on existing editors and
therefore cannot fall back on refined implementations of features that
have been modified and improved many times throughout their existence.

\pddlmode\footnote{http://rakaposhi.eas.asu.edu/planning-
  list-mailarchive/msg00085.html } for the widely used Emacs editor
builds on the sophisticated features of Emacs and uses its
extensibility and customizability. It provides syntax highlighting by
way of basic pattern matching of keywords, variables, and
comments. Additional features are automatic indentation and code
completion as well as bracket matching. Code snippets for the creation
of domains, problems, and actions are also available. Finally,
\textsc{pddl}-mode keeps track of action and problem declarations by
adding them to a menu and thus intending to allow for easy and fast
code navigation.

\pddlmode for Emacs supports \textsc{pddl} versions up to
2.2, which includes derived predicates and timed initial predicates
\cite{edelkamp2004pddl2}, but does not recognize later features like
object-fluents.

The tool editor.planning.domains allows for editing \textsc{pddl}
files online. Its feature comprise syntax highlighting, code folding,
\textsc{pddl}-specific auto-completion, and multi-tab support. The
editor is part of the Planning.Domains initiative which aims at
providing three pillars to the planning community: (1) an API for
existing planning problems; (2) a planner-in-the-cloud service; and
(3) an online PDDL editor. In contrast to \mypddl,
editor.planning.domains does not support context-aware syntax
highlighting: for example, editor.planning.domains highlights key
words even if they are misplaced. In contrast, \mypddl is aware of the
scope of a construct and highlights key words accordingly.

The tool
vscode-pddl\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
offers a wide range of editing functions, such as syntax highlighting,
code completion, and code snippets. Since it is implemented as a
plugin for the editor Visual Studio Code, it can fall back on the the
extensibility and maturity of this code editor. In contrast to
\mypddl, it does not offer domain visualization nor an interface with
a programming language in order to extend \pddl's limited modeling
capacities.

In sum, there is currently no tool available supporting all features
of \pddl 3.1, nor all the steps in the modeling process.

\section{MyPDDL}
\label{sec:mypddl}

\mypddl is a highly customizable and extensible modular system,
designed for supporting knowledge engineers in the process of writing,
analyzing and expanding \pddl files and thereby promoting the
collaboration between knowledge engineers and the use of \pddl in
real-world applications. \mypddl consists of the following integral
parts:

\subsection{Modules}

\begin{description}
\item[\mypddlide] is an integrated development environment (IDE) for
  the use of \mypddl in the text and code editor \emph{Sublime
    Text}\footnote{\url{http://www.sublimetext.com}}. Since
  \mypddlsnippet and \textsc{-syntax} are devised explicitly for
  \sublimetext, their integration is implicit. The other tools (-new,
  -diagram, -distance) can be used independently of \sublimetext with
  the command-line interface and any \pddl file, but were also
  integrated into the editor.

\item[\mypddlsyntax] is a context-aware syntax highlighting feature
  for \sublimetext. It distinguishes all \pddl constructs up to
  version 3.1. Using regular expressions that can recognize both the
  start and the end of code blocks by means of a sophisticated pattern
  matching heuristic, \mypddlsyntax identifies \pddl code blocks and
  constructs and divides them into so called scopes, i.e. named
  regions. \sublimetext colorizes the code elements via the assigned
  scope names and in accordance with the current color scheme. These
  scopes allow for a fragmentation of the \pddl files, so that
  constructs are only highlighted if they appear in the correct
  context. Thus missing brackets, misplaced expressions and misspelled
  keywords are visually distinct and can be identified (see
  Figure~\ref{fig:syntax}).

  \begin{figure}
    \centering
    \includegraphics[width=0.6\textwidth]{coffee-yes.pdf}
    \caption{Syntax highlighting using \mypddlide. White
      text contains errors.}
\label{fig:syntax}
  \end{figure}

\item[\mypddlnew] helps to organize \pddl projects by generating the
following folder structure:

\begin{figure}[h] 
  \dirtree{%
  .1 project-name/.
  .2 domains/.
  .2 problems/.
  .3 p01.pddl.
  .2 solutions/.
  .2 domain.pddl.
  .2 README.md.
  .2 plan.
  }
\end{figure}

The domain file \texttt{domain.pddl} and the problem file
\texttt{p01.pddl} initially contain corresponding \pddl skeletons
which can also be customized. All problem files that are associated
with one domain file are collected in the folder
\texttt{problems/}. \texttt{README.md} is a Markdown file, which is
intended for (but not limited to) information about the author(s) of
the project, contact information, informal domain and problem
specifications, and licensing information.  Markdown files can be
converted to \textsc{html} by various hosting services (like GitHub or
Bitbucket). The basic planner integration \mypddlplan provided by the
file \texttt{plan} is described below.

\item[\mypddlsnippet] provides code skeletons, i.e. templates for
  often used pddl constructs such as domains, problems, type and
  function declarations, and actions. They can be inserted by typing a
  triggering keyword. Almost all \pddl actions consist of these same parts. And this is only
an example. Knowledge engineers have to use the same constructs again
and again when writing and extending \pddl files. This is where code
snippets can come in handy. To facilitate and speed up the implementation of
standard constructs, \mypddl-snippet provides code skeletons, i.e.
templates for often used \pddl constructs. They can be inserted by
typing a triggering keyword (trigger). Table \ref{tab:snippets}
displays descriptions of all available snippets and the corresponding
trigger.

\begin{table}[htb]
\centering
\begin{tabular}{ll}
  \tableheadline{snippet description} & \tableheadline{trigger}                   \\
  \hline
  domain skeleton                     & \texttt{domain}                           \\
  problem skeleton                    & \texttt{problem}                          \\
  type declaration                    & \texttt{t1, t2, ...}                      \\
  typed predicate declaration         & \texttt{p1, p2, ...}                     \\ 
  typed function declaration          & \texttt{f1, f2, ...}                     \\
  action skeleton                     & \texttt{action}, \texttt{durative-action} \\
\end{tabular}\caption[Available snippets in \mypddl-snippet]{\label{tab:snippets}The snippets that can be inserted into \pddl files by typing the trigger.}

\end{table}

For example, typing \texttt{action} and pressing the tabulator key (this
function defaults to the tabulator key, but the key can be customized)
inserts the action skeleton in Listing \ref{ls:action-skeleton}. \pddl
constructs with a specified arity can be generated by adding the arity
number to the trigger (\texttt{p2} would insert the binary predicate
template \texttt{(pred-name~?x~-~object~?y~-~object)}.

Once the snippet has been inserted, skipping from blank to blank is
enabled by pressing the tabulator key. This allows for a fast
navigation within the snippet.

Every snippet is stored in a separate file, located in the folder
\texttt{Packages/PDDL/} of Sublime Text. New snippets can be added and
existing snippets can be customized (by changing the template or the
trigger) in this folder.

\item[\mypddlclojure] provides a preprocessor for \pddl files to
  bypass \pddl's limited mathematical capabilities, thus reducing
  modeling time without overcharging planning algorithms. Since \pddl
  is used to create more and more complex domains
  \cite{goldman2012type,guerin2012academic}, needing the square root
  function for a distance optimization problem or the logarithmic
  function for modeling an engineering problem does not seem to be a
  far-fetched scenario. However, \pddl's calculating capabilities are
  limited \cite{parkinson2012increasing}. While these mathematical
  operations are currently not supported by \pddl itself,
  preprocessing \pddl files in a programming language and then
  hardcoding the results back into the file seems to be a reasonable
  workaround. With the help of such an interface, the modeling time
  can reduced an even partly automated (see \ref{subsec:loc} the
  distance calculator \mypddl-distance).  We decided to use Clojure
  \cite{hickey2008clojure}, a modern Lisp dialect that runs on the
  Java Virtual Machine (\textsc{jvm}) \cite{lindholm2014java},
  facilitating input and output of the Lisp-style \pddl
  constructs. The interface is provided as a Clojure library and based
  on two methods:
\begin{description}
\item[{read-construct(keyword, file)}] Allows for the extraction of
  code blocks from \pddl files. Listing \ref{src:read-construct} shows
  an example in which the goal state of \emph{Gary's Huge Problem} is
  extracted.
\end{description}
\begin{description}
\item[{add-construct(file, block, part)}] Provides a means for adding
constructs to a specified code block in \pddl problem files. This
is illustrated in Listing \ref{ls:using-interface} and
\ref{ls:after-interface} where the predicate \texttt{(hungry gisela)} is
added to the \texttt{(:init~...)} block of \emph{Gary's Huge Problem}.
\end{description}

\begin{listing}[H]
\begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{clojure}
(read-construct :goal "garys-huge-problem.pddl")
;;=> ((:goal (exploited magicfailureapp)))
\end{minted}
\caption[Extracting the goal state using the interface]{\label{src:read-construct}Extracting the goal state of \emph{Gary's Huge Problem} by using the interface.}
\end{listing}

\begin{description}
\item[{add-construct(file, block, part)}] Provides a means for adding
constructs to a specified code block in \pddl problem files. This
is illustrated in Listing \ref{ls:using-interface} and
\ref{ls:after-interface} where the predicate \texttt{(hungry gisela)} is
added to the \texttt{(:init~...)} block of \emph{Gary's Huge Problem}.
\end{description}

\begin{listing}[H]
\begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{clojure}
(add-construct "garys-huge-problem.pddl" :init '((hungry gisela)))
\end{minted}
\caption[Adding a predicate using the
interface]{\label{ls:using-interface}Adding the predicate
  \texttt{(hungry~ gisela)} to the \texttt{(:init~...)} block of
  \emph{Gary's Huge Problem} using the PDDL/Clojure interface.}
\end{listing}

\begin{listing}[H]
  \begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
(:init (hungry gary)
       (in pizza-box big-pepperoni)
       (has-access gisela magicfailureapp))
       (hungry gisela))
\end{minted}
\caption[Modified problem file]{\label{ls:after-interface}The modified
  part of the problem file to which the predicate \texttt{(hungry gisela)} has
  been added.}
\end{listing}

\item[\mypddldistance] provides special preprocessing functions for
  distance calculations. Let us assume that Gary from Chapter \ref{ch:basics} wants the
\emph{Hacker World} to be more realistic. Every object shall now have
a location specified by a x and y coordinate. Imagine this domain now
including a further precondition in the action \texttt{eat-pizza}, so
that Gary can only eat the pizza if it is at most an arm's length
away. While the location of objects can be implemented using the
predicate \texttt{(location ?o - object ?x ?y - number)}, with
\texttt{x} and \texttt{y} being the spatial coordinates of an object,
calculating the Euclidean distance requires using the square root
function ($\sqrt{}$). Unfortunately, \pddl 3.1 supports only four
arithmetic operators (\texttt{+}, \texttt{-}, \texttt{/}, \texttt{*}).
How then is it possible to prevent hackers from starving?

\cite{parkinson2012increasing} describe a workaround for this
drawback. By writing an action \texttt{calculate-sqrt}, they bypass
the missing square root function by making use of the Babylonian root
method. Although this method approximates the square root function, it
requires many iterations and would most likely have an adverse effect
on plan generation \cite{parkinson2012increasing}.

More usable and probably faster results can be achieved by using the
interface between \pddl and Clojure as a distance calculator,
implemented in the tool \mypddl-distance. It reads a problem file into
Clojure and extracts all locations, defined in the \texttt{(:init
  ...)} code block. The Euclidean distances between these locations
are then calculated and written back into a new and now extended copy
of the problem file, using the predicate
\texttt{(distance~?o1~?o2~-~object ?n~-~number)}\footnote{This
  predicate has to be declared in the \texttt{(:predicates ...)} block
  of the corresponding domain file.}, which specifies the distance
between two objects. Listing \ref{ls:before-mypddl-distance} and
\ref{ls:after-mypddl-distance} show the \texttt{(:init ...)} code
block of a extended version of \emph{Gary's Huge Problem} before and
after using \mypddl-distance, respectively.

The calculator works on any arity of the specified location predicate,
so that locations can be specified one, two and three dimensionally
and even used in higher dimensions.

However, this preprocessing of problem file also has a drawback, apart
from the time required to calculate possibly unused distances. If the
number of locations is \(n\), the number of calculated distances is
\(n^2\), because every location has a distance to every other location
(and to itself). The calculated distances have to be stored in the
\pddl problem file, potentially requiring a lot of space. Therefore, a
sensible next step would be to extend \pddl by increasing its
mathematical expressivity \cite{parkinson2012increasing}, perhaps by
declaring a requirement \texttt{:math} that specifies further
mathematical operations.

For domains with spatial components, the distance of objects is often
important and should not be omitted in the domain model. However,
calculating distances from coordinates requires the square root
function, which is not supported by \pddl (it only supports the four
basic arithmetic operators). More sophisticated calculations can be
achieved with the supported operators, but the solutions are rather
inefficent and inelegant \cite{parkinson2012increasing}. By
calculating the distances offline and including them as additional
predicates in the problem file using \mypddldistance, the distances
between objects are given to the planner as part of the problem
description.

\begin{listing}[H]
\begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
(:init ...
       (location gary 4 2)
       (location pizza 2 3))
\end{minted}
\caption[Excerpt of the extended problem file before using
\mypddl-distance]{\label{ls:before-mypddl-distance}Excerpt of the
  \texttt{(:init~...)} block of the extended file \emph{Gary's Huge
    Problem} before using \mypddl-distance.}
\end{listing}

\begin{listing}[H]
\begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
(:init ...
       (location gary 4 2)
       (location pizza 2 3)
       (distance gary gary 0.0)
       (distance gary pizza 2.2361)
       (distance pizza gary 2.2361)
       (distance pizza pizza 0.0))
\end{minted}
\caption[Excerpt of the extended problem file after using
\mypddl-distance]{\label{ls:after-mypddl-distance}After the
  application of \mypddl-distance, the calculated distances are
  inserted in the \texttt{(:init ...)} code block in a copy of the
  problem file.}
\end{listing}

\item[\mypddldiagram] generates a \textsc{png} image from a \pddl
  domain file as shown in Figure~\ref{fig:diagram}. The diagrammatic
  representation of textual information helps to quickly understand
  the connection of hierarchically structured items and should thus be
  able to simplify the communication and collaboration between
  developers. In the process of generating the diagrams, an optional
  copy of the \pddl file can be created, providing a simple version
  control system. In the diagram, types are represented with
  boxes, with every box consisting of two parts:

  \begin{itemize}
  \item The header displays the name of the type.
  \item The lower part displays all predicates that use the
    corresponding type at least once as a parameter. The predicates
    are written just as they appear in the \pddl code.
\end{itemize}

Generalization relationships express that every subtype is also an
instance of the illustrated super type (e.g. ``a laptop \emph{is a}
computer''). This relationship is indicated in the diagram with an
arrow from the subtype (here: \emph{laptop}) to the super type (here:
\emph{computer}).

In order to create the diagram, \mypddl-diagram utilizes dot from the
Graphviz package\footnote{Graphviz is a collection of programs for
  drawing graphs. This includes dot, a scriptable, graphing tool, that
  is able to generate hierarchical drawings of directed graphs in a
  variety of output formats (e.g. \textsc{png}, \textsc{pdf},
  \textsc{svg}). The input to dot are text files, written in the
  \textsc{dot} language.} \cite{ellson2002graphviz} and takes the
following steps:

\begin{enumerate}
\item A copy of the domain file is stored in the folder
  \texttt{domains/}.
\item The \texttt{(:types~...)} block is extracted via the PDDL/Clojure
  interface.
\item In Clojure, the types are split into super types and associated
  subtypes using regular expressions and stored in a Clojure hashmap.
\item Based on the hashmap, the description of a directed graph in the
  \textsc{dot} language is created and saved in the folder
  \texttt{dot/}.
\item The \textsc{dot} file is passed to dot, creating a \textsc{png}
  diagram and saving it in the folder
  \texttt{diagrams/}.
\item The \textsc{png} diagram is displayed in a window.
\end{enumerate}

Every time \mypddl-diagram is invoked, these steps are executed and
the names of the saved files are extended by an ascending revision
number. Thus, one cannot only identify associated \pddl, \textsc{dot}
and \textsc{png} files, but also use this feature for basic revision
control. Figure~\ref{fig:mypddl-new-project-folder} displays the
folder structure after invoking \mypddl-diagram twice on the
\emph{Hacker World}.

  \begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{dot-diagram}
    \caption{Type diagram generated by \mypddldiagram}
    \label{fig:diagram}
  \end{figure}

\item[\mypddlplan] is a basic planner integration for \mypddl. After
  creating a new project with \mypddlnew, the file \texttt{plan} in
  the project folder contains a shell script for executing a planner
  with the new domain and problem files as input. The desired planner
  can be specified in the file \texttt{plan} or by editing the
  template. Due to the versatility of shell scripts, any planner can
  be used and arbitrary command line options can be specified.

\end{description}

\subsection{Design Principles}

As guidelines for design decisions, we used the seven criteria for
knowledge engineering tools proposed by Shah et
al. \cite{shah2013knowledge} as well as general usability
principles. Operationality instantiates, whether the generated models
can improve the planning performance. This is not a design principle
for \mypddl, because we assume that \mypddl does not improve the
quality (with respect to planning performance) of the resulting \pddl
specifications. Therefore, we replaced this criterion with functional
suitability from the \textsc{iso/iec} 25010 standard, which is defined
as “the degree to which the software product provides an appropriate
set of functions for specified tasks and user objectives”
(\textsc{iso} 25010 6.1.1). \mypddl supports the current version 3.1
of \pddl. It encompasses and exceeds most of the functionality of the
existing tools. It specifically supports basic editor features with a
high customizability as well as visualization support.

\begin{description}

\item[Collaboration:] With the growing importance of team work and
  team members not necessarily working in the same building, or in the
  same country, there is an increasing need for tools supporting the
  collaboration effort. In developing \mypddl, this need was sought to
  be met by \mypddldiagram. Complex type hierarchies can be hard to
  overlook, especially if they were constructed by someone
  else. Therefore, a good way of tackling this problem seemed to be by
  providing a means to visualize such hierarchies in the form of type
  diagrams.

\item[Experience:] \mypddl was designed specifically for users with a
  background in AI, but not necessarily in \pddl. The tools are
  similar to standard software engineering tools and should thus be
  easily learnable. The user evaluation (Section TODO) confirms that
  \mypddl helps novices in \pddl to master planning task modeling. In
  addition, it is also possible to customize \mypddl so as to adapt
  its look and feel to other programs one is already familiar with, or
  simply to make it more enjoyable to use. The project
  site\footnote{\url{http://pold87.github.io/myPDDL}} provides \mypddl
  video introductions and a manual to get started quickly.

\item[Efficiency:] All \mypddl tools are intended to increase the
  efficiency with which \pddl files are created. \mypddlsnippet
  enables the fast creation of large and correct code skeletons that
  only need to be complemented. \mypddlsyntax can reduce the time
  spent on searching errors. Code folding allows users to hide
  currently irrelevant parts of the code and automatic indentation
  increases its readability. To easily keep track of all the parts of
  a project, folders are automatically created and named with
  \mypddlnew. \mypddlclojure and \textsc{-distance} allow
  for a straightforward inclusion of numerical values in the problem
  definition.

\item[Debugging:] \mypddl\textsc{-syntax} highlights all syntactically
  correct constructs and leaves all syntactical errors
  non-highlighted. In contrast, \pddlmode for Emacs and itSimple only
  provide basic syntax highlighting for emphasizing the
  structure. \pddlstudio explicitly detects errors, but the user is
  immediately prompted when an error is detected. Often, such error
  messages are premature, for example, just because the closing
  parenthesis was not typed yet, does not mean it was
  forgotten. \mypddl indicates errors in a more subtle way: syntactic
  errors are simply not highlighted, while all correct \pddl code
  is. The colors are customizable, so that users can choose how
  prominent the highlighting sticks out.

\item[Maintenance:] The possibility to maintain \pddl files is a key
  aspect of \mypddl. The automatically generated type diagram
  (\mypddldiagram) gives an overview of the domain structure and
  thereby serves as a continuous means of documentation. Helping to
  understand foreign code, though, it follows logically that
  \mypddldiagram also helps in coming back and changing one’s own
  models if some time has elapsed since they were last edited. The
  basic revision control feature of \mypddldiagram keeps track of
  changes, making it easy to revert to a previous domain
  version. Furthermore, \mypddlnew encourages adhering to an organized
  project structure and stores corresponding files at the same
  location.  The automatically created readme file can induce the user
  to provide further information and documentation about the \pddl
  project.

\item[Support:] \mypddlide can be installed using \sublimetext's
  Package
  Control\footnote{\url{https://sublime.wbond.net/about}}. This allows
  for an easy installation and staying up-to-date with future
  versions. In order to provide global access and with it the
  possibility for developing an active community, the project source
  code is hosted on
  GitHub\footnote{https://github.com/Pold87/myPDDL}. Additionally, the
  project site provides room for discussing features and reporting
  bugs.

\end{description}

\section{Validation and Evaluation}
\label{sec:valid-eval}

To assess the utility of \mypddl, we used the criteria listed in
Section TODO. The functional suitability was evaluated using a
benchmark validation, comparing \mypddl's functionality with the tools
described in Section 2. The criteria collaboration, experience,
efficiency, and debugging were evaluated in a user test. We analyzed
the user performance both with and without using \mypddlsyntax and
\mypddldiagram. The \mypddl components supporting maintenance are the
same ones that are used in the user test, but their long-term usage is
difficult to evaluate. The support criterion depends primarily on the
infrastructure, which has been established as explained in 3.2.

\subsection{Benchmark Validation}

Functional suitability encompasses the set of functions to meet the
user objectives. The tools of Section 2 basically all follow the same
objectives as \mypddl: creating \pddl domains and problems. The
features offered by each tool are summarized in Table
\ref{tool-comp}. Besides supporting the latest \pddl version, a
strength of \mypddl is its high customizability, which comes with the
\sublimetext editor. Being the only one of the six tools capable of
visualizing parts of the \pddl code, it must be understood as
complementary to \itsimple, which takes the opposite approach of
transforming \uml diagrams into \pddl files. The fact that \mypddl
does not check for semantic errors is not actually a drawback because
planners usually detect semantic errors. All in all, \mypddl strives
to support the planning task engineer during all phases of the
modeling process. Additionally, it features some unique tools, such as
domain visualization and an interface with a programming language. It
can therefore be concluded that \mypddl provides an appropriate set of
functions for developing \pddl files and is thus functionally
suitable.

\begin{sidewaystable}[]
\centering
\footnotesize
\begin{tabularx}{\textwidth}{lX|llll}
  \tableheadline{Feature}             & \tableheadline{Function}                              & \tableheadline{\pddlstudio} & \tableheadline{\itsimple} & \tableheadline{\pddlmode} & \tableheadline{editor.planning.domains} & \tableheadline{PDDL - Visual Studio} & \tableheadline{\mypddl} \\
  \hline
  latest supported \pddl version      & considering recent \pddl features                     & 1.2                         & 3.1                       & 2.2                           &   3.1    &  3.1    & 3.1       \\
  syntax highlighting                 & supporting error detection and code navigation        & yes                         & basic                     & basic                         &   basic  & yes     & yes       \\
  semantic error detection            & supporting error detection                            & yes                         & no                        & no                            &   no     & yes     & no        \\
  automatic indentation               & supporting readability and navigation                 & no                          & no                        & yes                           &   no     & yes     & yes       \\
  code completion                     & speeding-up the knowledge engineering process         & yes                         & no                        & yes                           &   yes    & yes     & yes       \\
  code snippets                       & speeding-up the knowledge engineering process         & no                          & yes                       & yes                           &   basic  & yes     & yes       \\
                                      & externalizing user's memory                           &                             &                           &                               &          &         &           \\
  code folding                        & supporting keeping an overview of the code structure  & yes                         & no                        & yes                           &   yes    & yes     & yes       \\
  domain visualization                & supporting fast understanding of the domain structure & no                          & no                        & no                            &   no     & no      & yes       \\
  project management                  & supporting keeping an overview of associated files    & yes                         & yes                       & no                            &   yes    & yes     & yes       \\
  \uml to \pddl code translation      & supporting initial modeling                           & no                          & yes                       & no                            &   no     & no      & yes       \\
  planner integration                 & allowing for convenient planner access                & basic                       & yes                       & no                            &   yes    & yes     & yes       \\
  plan visualization                  & supporting understanding and crosschecking the plan   & no                          & yes                       & no                            &   no     & yes     & no        \\
  dynamic analysis                    & supporting dynamic domain analysis                    & no                          & yes                       & no                            &   no     & no      & no        \\
  declaration menu                    & supporting code navigation                            & no                          & no                        & yes                           &   no     & yes     & no        \\
  interface with programming language & automating tasks                                      & no                          & no                        & no                            &   no     & no      & yes       \\
                                      & extending \pddl's modeling capabilities               &                             &                           &                               &          &      &              \\
  customization features              & acknowledging individual needs and preferences        & basic                       & no                        & yes                           &  basic   & yes     & yes       \\
\end{tabularx}\caption[Comparison of knowledge engineering
tools]{\label{tool-comp}Comparison of knowledge engineering tools and their features.}

\end{sidewaystable}

\subsection{User Evaluation}

The two most central modules of \mypddl are \mypddlsyntax and
\mypddldiagram, since they support collaboration, efficiency, and
debugging independently of the user’s experience with \pddl. To
evaluate their usability, they were evaluated in a user study. To this
end, we compared the user performance regarding several tasks, both
with and without using the respective module.

\subsubsection{Participants}

In Usability Engineering, a typical number of participants for user
tests is five to ten. Studies have shown that even such small sample
sizes identify about 80\,\% of the usability problems
\cite{nielsen1994estimating,hwang2010number}. Our study design
required eight participants. Three female and five male participants
took part in the study (average age of 22.9, standard deviation of age
0.6). All participants were required to have basic experience with at
least one Lisp dialect (in order not to be confused with the many
parentheses), but no experience with \pddl or AI planning in general.

\subsubsection{Approach}

No earlier than 24 hours before the experiment was to take place,
participants received the web
link\footnote{\url{https://www.youtube.com/watch?v=Uck-K8VnNOU&list=PL3CZzLUZuiIMWEfJxy-G6OxYVzUrvjwuV} (tutorial is in German)}
to a 30-minute interactive video tutorial on AI planning and
\pddl. This method was chosen in order not to pressure the participant
with the presence of an experimenter when trying to understand the
material.

\subsubsection{Procedure}

We defined four tasks: two debugging tasks for testing the syntax
highlighting feature and two type hierarchy tasks for testing the type
diagram generator. A within subjects design was considered most suited
dueto the small number of participants. Therefore, it was necessary to
construct two tasks (matched in difficulty) for each of these two
types to compare the effects of having the tools available. Each
participant started either with a debugging or type hierarchy task and
was given the \mypddl tools either in the first two tasks or the
second two tasks, so that each participant completed each task type
once with and once without \mypddl. This results in $2$ (first task is
debugging or hierarchy) $\times$ $2$ (task variations for debugging
and hierarchy) $\times$ $2$ (starting with or without \mypddl) $= 8$
individual task orders, one per participant.

\begin{itemize}
\item Debugging Tasks

  For the debugging tasks, participants were given six minutes (a
  reasonable time frame tested on two pilot tests) to detect as many
  of the errors in the given domain as possible. They were asked to
  record each error in a table using pen and paper with the line
  number and a short comment. Moreover, they were instructed to
  immediately correct the errors in the code if they knew how to, but
  not to dwell on the correction otherwise. For the type hierarchy
  task, participants were asked to answer five questions concerning
  the domains, all of which could be facilitated with the type diagram
  generator. One of the five questions (Question 4) also required
  looking into the code. Participants were told that they should not
  feel pressured to answer quickly, but to not waste time either. Also
  they were asked to say their answer out loud as soon as it became
  evident to them. They were not told that the time it took them to
  come up with an answer was recorded, since this could have made them
  feel pressured and thus led to more false answers.

\item Type Hierarchy Tasks

  The two tasks to test syntax highlighting presented the user with
  domains that were 54 lines in length, consisted of 1605 characters
  and contained 17 errors each. Errors were distributed evenly
  throughout the domains and were categorized into different
  types. The occurrence frequencies of these types were matched across
  domains as well, to ensure equal difficulty for both domains. To
  test the type diagram generator, two fictional domains with equally
  complex type hierarchies consisting of non-words were designed (five
  and six layers in depth, 20 and 21 types). The domains were also
  matched in length and overall complexity (five and six predicates
  with approximately the same distribution of arities, one action with
  four predicates in the precondition and two and three predicates in
  the effect).

\item System Usability Scale

  At the end of the usability test the participants were asked to
  evaluate the perceived usability of \mypddl using the system
  usability scale \cite{brooke1996sus}.

\end{itemize}

\subsubsection{Analysis}

\begin{itemize}
\item Debugging Tasks
\nopagebreak
  To compare differences in the debugging tasks, a paired sample
  $t$-test was used; normality was tested with a Shapiro-Wilk
    test. to compare the arithmetic means ($M$s) of detected
  errors. The test was performed two-tailed, since syntax highlighting
  might both help or hinder the participants. Arithmetic standard
  deviations ($SD$s) were calculated for each condition.

\item Type Hierarchy Tasks
\nopagebreak
  For the type hierarchy tasks, $t$-tests were performed on the
  logarithms of the data values to compare the geometric means for the
  two conditions for each question; normality was tested with a
  Shapiro-Wilk test on the log-normalized data values. The geometric
  mean is a more accurate measure of the mean for small sample sizes
  as task times have a strong tendency to be positively skewed
  \cite{sauro2012quantifying}. The geometric standard deviation
  ($GSD$) was calculated for each question and condition. Only those
  task completion times were included in the calculation of the
  $t$-values, where the respective participant gave a correct answer
  for both occurrences of a question. This approach should reduce the
  influence of random guessing. Again, two-tailed $t$-tests were used
  to account for both, improvements and drawbacks, of using
  \mypddldiagram.

\item System Usability Scale

  The arithmetic mean and standard deviation for the score on the
  System Usability Scale was calculated.
\end{itemize}

\subsubsection{Results}

\begin{itemize}
\item Debugging Tasks

  The participants detected more errors using the syntax highlighting
  feature, ($M = 7.6$, $SD = 2.07$) than without it ($M = 10.3$,
  $SD = 3.45$); $t(7) = 2.68$, $p = 0.03$.  That is, approximately
  36\,\% more errors were found with syntax highlighting. The
  arithmetic means are displayed in Figure
  \ref{fig:found-errors-combined}, where each cross~($\times$)
  represents the data value of one participant.
\begin{figure}[h]
  \centering
  \hspace{1.7cm}
  \includegraphics[width=0.7\textwidth]{found_errors}
  \caption[Diagram of detected errors]{Comparison of detected errors
    with and without the syntax highlighting feature. Each cross
    ($\times$) shows the data value of one participant. The bars
    display the arithmetic mean.}
\label{fig:found-errors-combined}
\end{figure}

\item Type Hierarchy Tasks

  Figure~\ref{fig:task-completions-agg} shows the geometric mean of
  the completion time of successful tasks for each question with and
  without the type diagram generator. With the type diagram generator
  participants answered all questions (except Question 4) on average
  nearly twice as fast ($GM = 33.0$, $GSD = 2.23$) as without it
  ($GM = 57.8$, $GSD = 2.05$); $t(32) = -3.34$, $p = 0.002$. This
  difference slightly increases, if Question 4 is excluded from the
  calculations: with type generator: $GM = 31.1$, $GSD = 2.17$,
  without: $GM = 58.1$, $GSD = 2.07$; $t(30) = -3.68$, $p < 0.001$.
  Table~\ref{tab:p-values} gives an overview of geometric means,
  geometric standard deviations, $t$-values, and $p$-values for each
  question.

  \begin{table}[]
    \centering
    \begin{tabular}{rrrrrrrr}
    \toprule
         & \multicolumn{4}{c}{Type Diagram Generator} & & &                                                                   \\
         & \multicolumn{2}{c}{with}                   & \multicolumn{2}{c}{without} &      &       &                     \\
\cmidrule(r){2-3}\cmidrule(r){4-5}
Question & $GM$                                       & $GSD$                       & $GM$ & $GSD$ & $df$ & $t$   & $p$  \\
\midrule
      Q1 & 21.8                                       & 1.52                        & 40.0 & 2.26  & 7    & -1.86 & 0.11 \\
      Q2 & 23.8                                       & 1.49                        & 50.8 & 2.16  & 7    & -1.91 & 0.10 \\
      Q3 & 48.0                                       & 3.49                        & 83.2 & 2.20  & 5    & -0.86 & 0.43 \\
      Q4 & 84.3                                       & 2.22                        & 54.1 & 1.93  & 1    & 4.48  & 0.14 \\
      Q5 & 41.2                                       & 2.24                        & 78.0 & 1.48  & 7    & -2.75 & 0.03 \\
\bottomrule
    \end{tabular}

    \caption{Overview of geometric means ($GM$s), geometric standard deviations ($GSD$s), degrees of freedom ($df$), $t$-values, and $p$-values. Please note, that the calculation for Q4 is based on only two paired data values ($df = 1$). Please note further that this table only considers paired data values, this means only if a participant answered the question correctly in both domains, the data value is considered (since \emph{paired} $t$-tests are calculated). In contrast, Figure \ref{fig:task-completions-agg} displays the geometric means for all correct answers.}

    \label{tab:p-values}
  \end{table}

\begin{figure}[h]
\captionsetup{singlelinecheck=off}
  \centering
  \hspace{1.7cm}
  \includegraphics[width=0.85\textwidth]{task_time}
  \caption[Diagram of the task completion time]{Task completion time
    for the type hierarchy tasks. The bars display the geometric mean
    averaged over all participants; each cross~($\times$) represents
    the data value of one participant. The percent values at the
    bottom of the bars show the percentage of users that completed the
    task successfully. The question can be found in the Appendix
    (\ref{splisus}, \ref{store}).}\label{fig:task-completions-agg}
\end{figure}

\item System Usability Scale
  \mypddl reached a score of 89.6 on the system usability scale
  \cite{brooke1996sus}, with a standard deviation of 3.9. 
\end{itemize}

\subsubsection{Discussion}

The user test shows that \mypddlsyntax and -\textsc{diagram} provide useful
tools for novices in AI planning and \pddl. Below, we will discuss
each part of the user test in turn.

\begin{itemize}
\item Debugging Tasks

  While, in general, the syntax highlighting feature was considered
  very useful, two participants remarked that the used colors confused
  them and that they found them more distracting than helpful. One of
  them mentioned that the contrast of the colors was so low that they
  were hard for her to distinguish. She found the same number of
  errors with and without syntax highlighting. The other of the two
  was the only participant who found less errors with syntax
  highlighting than without it. With \mypddlsyntax, two participants
  found all errors in the domain, while none achieved this without
  syntax highlighting.

\item Type Hierarchy Tasks

  In spite of the rather large difference between the $GM$s for
  Question 3, a high $p$-value is obtained ($p = 0.43$). This might be
  due to the high $GSD$ for the \emph{with} condition and the rather
  small degrees of freedom ($df = 5$). Testing more participants would
  probably yield clearer results here. The fact that the availability
  of tools did not have a positive effect on task completion times for
  Question 4 can probably be attributed to the complexity of this
  question (see Appendix, \ref{splisus} and \ref{store}): in contrast
  to the other four questions, here, participants were required to
  look at the actions in the domain file in addition to the type
  diagram. Most participants were confused by this, because they had
  assumed that once having the type diagram available, it alone would
  suffice to answer all questions. This initial confusion cost some
  time, thus negatively influencing the time on the task.

  Visualization tools such as \mypddl -diagram can improve the
  understanding of unknown \pddl code and thus support
  collaboration. But users may be unaware of the limitations of such
  tools. A possible solution is to extend \mypddl -diagram to display
  actions, but this can overload the diagram and, especially for large
  domains, render it unreadable. Different views for different aspects
  of the domain or dynamically displayed content could integrate more
  data, but this also hides functionality, which is generally
  undesired for usability \cite{norman2002design}.

\item Sytem Usability Scale

Since the overall mean score of the system usability scale has an
approximate value of 68 with a standard deviation of 12.5
\cite{sauro2011practical}, the score of \mypddl is well above average
with a small standard deviation. A score of 89.6 is usually attributed
to superior products \cite{bangor2008empirical}. Furthermore, 89.6
corresponds approximately to a percentile rank of 99.8\,\%, meaning
that it has a better perceived ease-of-use than 99.8\,\% of the
products in the database used by Sauro \cite{sauro2011practical}.

\end{itemize}

In summary, the user test shows that customizability is important, as
not all users prefer the same colors or syntax highlighting at all and
their personal preferences seem to correlate with the effectiveness of
the tools.

\section{Conclusion}
\label{sec:conclusion}

We designed \mypddl to support knowledge engineers in all steps of
developing a planning domain. Additionally, it support engineers to
understand, modify, and extend existing \pddl planning domains.

This was realized with a set of tools comprising code editing
features, namely syntax highlighting and code snippets, a type diagram
generator, and a distance calculator. To also have all tools
accessible from one place, they were made available in the Sublime
Text editor. The different needs and requirements of knowledge
engineers are met by the modular, extensible and customizable
architecture of the toolkit and Sublime Text. The evaluation of
\mypddl has shown some initial evidence that it allows a faster
understanding of the domain structure, which could be beneficial for
the maintenance and application of existing task specifications and
for the communication between engineers. Users perceive it as easy and
enjoyable to use, and the increase in their performance when using
\mypddl underpins their subjective impressions. It is hoped that the
planning community not only accepts but actually uses \mypddl
intensely, leading to extensive feedback and continuous improvements.

In future work, \mypddl's set of features could be extended in several
directions.  The interface between \pddl and Clojure offers a basis
for creating dynamic planning scenarios. Applications could be the modeling of
learning and forgetting (by adding facts to or retracting facts from a
\pddl file) or the modeling of an ever changing real world via dynamic
predicate lists. Another way of putting the interface to use would be
by making the planning process more interactive, allowing for the
online interception of planning software in order to account for the
needs and wishes of the end user.

All in all, the overall increase of efficiency due to facilitated
collaboration and support in maintaining an overview should encourage
a shift of focus toward real world problems in knowledge
engineering. The full modeling potential can only be reached with
appropriate tools, with \mypddl hopefully leading to a broader
acceptance and use of \pddl for planning problems.

\bibliographystyle{splncs04}
\bibliography{mybibliography}

\end{document}

### 2019-07-03 paper.tex
upd

@@
 \definecolor{dhscodebg}{rgb}{0.95,0.95,0.95} % listing background
+\DeclareUnicodeCharacter{2212}{-}
 
@@
 \renewcommand\UrlFont{\color{blue}\rmfamily}
+\usepackage{csquotes}
 
@@
 \newcommand{\pddlstudio}{\textsc{pddl studio}\xspace}
+\newcommand{\epd}{editor.planning.domains\xspace}
 \newcommand{\itsimple}{\textsc{itSimple}\xspace}
@@
 
-\keywords{PDDL \and Planning}
+\keywords{PDDL \and Planning \and Knowledge Engineering}
 \end{abstract}
-%
-%
 
@@
 \cite{shah2013knowledge,keps2014}. To ensure a standardized modeling
-format, the Planning Domain Definition Language (\textsc{pddl})
+format, \pddl (\emph{Planning Domain Definition Language})
 \cite{mcdermott1998pddl} was developed and has become the de facto
@@
 
-The specification of Artificial Intelligence (AI) planning can be
-time-consuming and cubersome. While there is a large spectrum of
-different AI languages, the de facto standard language is PDDL
-(\emph{Planning Domain Definition Language}). However, so far its use
-is predominantly confined to academic labs. One of the reasons seems
-to be the missing support of knowledge engieering tools so simplify
-the task of specifying complex planning domains.
-
 In this chapter, we describe \mypddl
 (Figure~\ref{fig:mypddl-overview}), a knowledge engineering toolkit
-that supports the AI knowledge engineer in the development of complex
-AI planning problems. \mypddl is intended to support knowledge
-engineers throughout the entire design cycle of specifying planning
-tasks. In the initial stages, it allows for the creation of structured
-\textsc{pddl} projects that should encourage a disciplined design
-process. With the help of snippets, i.e. code templates, often used
-constructs can be inserted in \textsc{pddl} files. A syntax
-highlighting feature that speeds up the error-detection can come in
-handy in intermediate stages. Understanding the textual representation
-of complex type hierarchies in domain files can be confusing, so an
-additional tool enables their visualization. \textsc{pddl}'s limited
-modeling capabilities were bypassed by developing an interface that
-converts \pddl code into Clojure \cite{hickey2008clojure} code and
-vice versa.  Within this project, the interface was employed for a
-feature that calculates distances between objects specified in a
-problem model, but the interface provides numerous other possibilities
-and could also be used to further automate the modeling process. All
-of the features were integrated into the customizable and extensible
-Sublime Text \cite{sublimetext3} editor. Since the main aim in the
-development of the toolkit was for it to be easy to use and maintain,
-it is evaluated with regard to these criteria. The usability was
-assessed by means of a user test with eight subjects that had no prior
-experience with artificial intelligence planning. The results indicate
-that both error-detection and the understanding of a given domain can
-be facilitated by \mypddl.
-
-\begin{figure}[h]
+that supports the knowledge engineer in the entire design cycle of
+specifying planning tasks. In the initial stages, it allows for the
+creation of structured \pddl projects that should encourage a
+disciplined design process. With the help of snippets, that is, code
+templates, often used syntactic constructs can be inserted into \pddl
+files. A syntax highlighting feature that speeds up the
+error-detection can come in handy in intermediate
+stages. Understanding the textual representation of complex type
+hierarchies in domain files can be confusing, so an additional tool
+enables their visualization. \pddl's limited modeling capabilities
+were bypassed by developing an interface that converts \pddl code into
+code of the functional programming language Clojure
+\cite{hickey2008clojure} and vice versa.  Within this project, the
+interface was employed for a feature that calculates distances between
+objects specified in a problem model, but the interface provides
+numerous other possibilities and could also be used to further
+automate the modeling process. All of the features were integrated
+into the customizable and extensible Sublime Text \cite{sublimetext3}
+editor. Since the main aim in the development of the toolkit was for
+it to be easy to use and maintain, it is evaluated with regard to
+these criteria. The usability was assessed by means of a user test
+with eight subjects that had no prior experience with AI planning. The
+results indicate that both error-detection and the understanding of a
+given domain can be facilitated by \mypddl. This chapter is an
+extended version of work published in a previous paper~\cite{StrKir2014:aai}.
+
+\begin{figure}[H]
   \centering
-  \begin{tikzpicture}[
-  level 1/.style={sibling distance=40mm},
-  edge from parent/.style={->,draw},
-  >=latex]
-
-  % root of the the initial tree, level 1
-  \node[] (ide) {-IDE};
-\node[below of = ide,root] {myPDDL}
-% The first level, as children of the initial tree
-  child {node[level 2] (c1) {-new}}
-  child {node[level 2] (c2) {-syntax}}
-  child {node[level 2] (c3) {-snippet}}
-  child {node[level 2] (c4) {-distance}}
-  child {node[level 2] (c5) {-diagram}};
-
-% The second level, relatively positioned nodes
-\begin{scope}[every node/.style={level 3}]
-\node [below of = c1, xshift=15pt] (c11) {Setting shape};
-\node [below of = c11] (c12) {Choosing color};
-\node [below of = c12] (c13) {Adding shading};
-
-\node [below of = c2, xshift=15pt] (c21) {Using a Matrix};
-\node [below of = c21] (c22) {Relatively};
-\node [below of = c22] (c23) {Absolutely};
-\node [below of = c23] (c24) {Using overlays};
-
-\node [below of = c3, xshift=15pt] (c31) {Default arrows};
-\node [below of = c31] (c32) {Arrow library};
-\node [below of = c32] (c33) {Resizing tips};
-\node [below of = c33] (c34) {Shortening};
-\node [below of = c34] (c35) {Bending};
-\end{scope}
-
-% lines from each level 1 node to every one of its "children"
-\foreach \value in {1,2,3}
-  \draw[->] (c1.195) |- (c1\value.west);
-
-\foreach \value in {1,...,4}
-  \draw[->] (c2.195) |- (c2\value.west);
-
-\foreach \value in {1,...,5}
-  \draw[->] (c3.195) |- (c3\value.west);
+  \tikzstyle{every node}=[draw=black,anchor=west,
+  align=left,
+  scale=0.7]
+  \begin{tikzpicture}[%
+  grow via three points={one child at (0.5,-0.7) and
+    two children at (0.5,-0.7) and (0.5,-1.4)},
+  edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
+  \node {~~~\textbf{myPDDL}~~~}
+    child { node {\textbf{-new}\\Create new \pddl project}}		
+    child { node {\textbf{-syntax}\\Context-aware syntax highlighting}}
+    child { node {\textbf{-snippet}\\Code templates}}
+    child { node {\textbf{-clojure}\\Preprocessor for \pddl files via Clojure}}
+    child { node {\textbf{-distance}\\Distance calculation in spatial domains}}
+    child { node {\textbf{-diagram}\\Visualization of type hierarchy}}
+    child { node {\textbf{-plan}\\Planner integration}};
   \end{tikzpicture}
@@
 The remainder of this chapter is structured as
-follows. Section~\ref{sec:related-work} compares other PDDL editors
-and \pddl knowledge engineering tools. Section~\ref{sec:myPDDL}
-describes the different modules of \mypddl and their design
-principles. Section~\ref{sec:valid-eval} compares \mypddl to the other
-existing tools using a benchmark
-validation. Section~\ref{sec:conclusion} concludes the chapter and
-outlines future work.
+follows. Section~\ref{sec:related-work} compares \pddl knowledge
+engineering tools to lay a foundation for
+\mypddl. Section~\ref{sec:mypddl} describes the different modules of
+\mypddl and their design principles. Section~\ref{sec:valid-eval}
+evaluates \mypddl via a user test. Section~\ref{sec:conclusion}
+concludes the chapter and outlines future work.
 
@@
 
-This section introduces knowledge engineering tools that allow editing
-\textsc{pddl} files in a textual environment. After introducing the
-tools, they are compared and their shortcomings are discussed to set
-the stage for \mypddl.
-
-\pddlstudio \cite{plch2012inspect} is an application for creating and
-managing \pddl projects, i.e. a collection of \pddl
-files. \pddlstudio 's integrated development environment (\ide) was
-inspired by Microsoft Visual Studio and imperative programming
-paradigms. Its main features are syntax highlighting, error detection,
+This section introduces, compares, and discusses knowledge engineering
+tools that allow text-based editing of \textsc{pddl} files to set the
+stage for \mypddl (Table~\ref{tab:comparison}).
+
+\pddlstudio \cite{chomuttool,plch2012inspect} is an IDE
+(\emph{integrated development environment}) for creating and managing
+\textsc{pddl} projects, that is, a collection of \textsc{pddl}
+files. Its main features are syntax highlighting, error detection,
 context sensitive code completion, code folding, project management,
-and planner integration. \pddlstudio 's error detection can recognize
-both syntactic (missing keywords, parentheses, etc.) and semantic
-(wrong type of predicate parameters, misspelled predicates, etc.)
-errors.
-
-A major drawback of \pddlstudio is that it is not updated regularly
-and only supports \pddl 1.2. Later \pddl versions contain several
-additional features such as durative actions, numeric fluents, and
-plan metrics \cite{edelkamp2004pddl2}.
-
-\itsimple \cite{vaquero2005itsimple} follows a graphical approach
-using Unified Modeling Language (\uml) diagrams. In the process
-leading up to \itsimple, \textsc{uml.p} (\uml in a Planning Approach)
-was proposed, a \uml variant specifically designed for modeling
-planning domains and problems \cite{vaquero2006use}.
-
-\itsimple's modeling workflow is unidirectional as changes in the
-\pddl domain do not affect the \uml model and \uml models have to be
-modeled manually, meaning that they cannot by generated from
-\pddl. However, \cite{tonidandel2006reading} present a translation
-process from a \pddl domain specification to an object-oriented
-\textsc{uml.p} model as a possible integration for \itsimple. This
-translation process makes extensive semantic assumptions for \pddl
-descriptions. For example, the first parameter in the
-\texttt{:parameters} section of an action is automatically declared as
-a subclass of the default class \texttt{Agent}, and the method is
-limited to predicates with a maximum arity of two. The current
-version of \itsimple does not include the translation process from
-\pddl to \uml.
+and planner integration. Many of these features are based on a parser,
+which continuously analyzes the code and divides it into syntactic
+elements. These elements and the way in which they relate to each
+other can then be identified. The syntax highlighter is a tool that
+colors constructs according to their syntactical meaning within the
+code. In the case of \pddlstudio, it colors names, variables, errors,
+keywords, predicates, types, and brackets each in a different
+customizable color.  \pddlstudio's error detection can recognize both
+syntactic (missing keywords, parentheses, etc.) and semantic (wrong
+type of predicate parameters, misspelled predicates, etc.)
+errors. This means that \pddlstudio can detect errors due to a
+mismatch between domain and problem file in real time. The code
+completion feature offers recommendations for standard standard
+\textsc{pddl} constructs as well as for previously used terms. Code
+folding allows the knowledge engineer to hide currently not needed
+code blocks. In this case only the first line of the block is
+displayed. All these above mentioned features of \pddlstudio utilize
+the parser. To keep track of all files, to display them in a tree
+structure, and to save them upon compilation, a project manager is
+integrated. Furthermore, this feature is necessary to detect interfile
+errors and to use the code completion functionality. Lastly, a
+command-line interface allows the integration of planners in order to
+run and compare different planning software.
+
+Unlike \pddlstudio, which provides a text based editor for \pddl, the
+\itsimple \cite{vaquero2005itsimple} editor has, as its main feature,
+a graphical approach that allows designing planning tasks in an
+object-oriented approach using \uml (\emph{Unified Modeling
+  Language}). In the process leading up to \itsimple,
+\textsc{uml.p}---\textsc{uml} in a Planning Approach---was proposed, a
+\textsc{uml} variant specifically designed for modeling planning
+domains and problems \cite{vaquero2006use}.
+
+The main purpose of \itsimple is supporting knowledge
+engineers in the initial stages of the design phase by making tools
+available that help transform the informality of the real world to
+formal specifications of domain models. The professed aim of the
+project is to provide a means to a \enquote{disciplined process of
+  elicitation, organization and analysis of requirements}
+\cite{vaquero2005itsimple}. However, subsequent design stages are also
+supported. Once domain and problem models have been created,
+\textsc{pddl} representations can be generated from the \textsc{uml.p}
+diagrams, edited, and then used as input to a number of different
+integrated planning systems. Compared to \pddlstudio, \itsimple has a
+different approach to planner integration: domain and problem can be
+fed to the planner by pressing a button, whereas in \pddlstudio, the
+user has to know and input commands in a command-line interface.
+
+With \itsimple, it is possible to directly input the domains and
+problems into a planner and to inspect the output from the planning
+system using the built-in plan analysis. This consists of a plan
+visualization, which shows the interaction between the plan and the
+domain by highlighting every change caused by an action. \itsimple's
+modeling workflow is unidirectional as changes in the \pddl domain do
+not affect the \uml model and \uml models have to be modeled manually,
+meaning that they cannot by generated using \pddl. Starting in version
+4.0 \cite{vaquero2012itsimple4} \itsimple expanded its features to
+allow the creation of \pddl projects from scratch (i.e. without the
+\uml to \pddl translation process). Thus far, the \pddl editing
+features are basic. A minimal syntax highlighting feature recognizes
+\pddl keywords, variables, and comments. Furthermore, \itsimple
+provides templates for \pddl constructs, such as requirement
+specifications, predicates, actions, initial, and goal definitions.
 
 Starting in version 4.0, \itsimple expanded its features to allow the
-creation of \pddl projects from scratch (i.e. without the \uml to
-\pddl translation process) \cite{vaquero2012itsimple4}. Thus far, the
-\pddl editing features are basic. A minimal syntax highlighting
-feature recognizes \pddl keywords, variables, and comments. \itsimple
-also provides templates for \pddl constructs, such as requirement
-specifications, predicates, actions, initial state, and goal
-definitions.
-
-Both \pddlstudio and \itsimple do not build on existing editors and
-therefore cannot fall back on refined implementations of features that
-have been modified and improved many times throughout their existence.
-
-\pddlmode\footnote{http://rakaposhi.eas.asu.edu/planning-
-  list-mailarchive/msg00085.html } for the widely used Emacs editor
-builds on the sophisticated features of Emacs and uses its
-extensibility and customizability. It provides syntax highlighting by
-way of basic pattern matching of keywords, variables, and
-comments. Additional features are automatic indentation and code
+creation of \pddl projects from scratch, that is, without the \uml to
+\pddl translation process \cite{vaquero2012itsimple4}. So far, a basic
+syntax highlighting feature recognizes \pddl keywords, variables, and
+comments. \itsimple also provides templates for \pddl constructs, such
+as requirement specifications, predicates, actions, initial state, and
+goal definitions.
+
+\pddlmode\footnote{\url{http://rakaposhi.eas.asu.edu/planning-
+    list-mailarchive/msg00085.html}} for Emacs builds on the
+sophisticated features of the widely used Emacs editor and uses its
+extensibility and customizability. \pddlmode provides syntax
+highlighting by way of basic pattern matching of keywords, variables,
+and comments. Additional features are automatic indentation and code
 completion as well as bracket matching. Code snippets for the creation
@@
 
-The tool editor.planning.domains allows for editing \textsc{pddl}
-files online. Its feature comprise syntax highlighting, code folding,
-\textsc{pddl}-specific auto-completion, and multi-tab support. The
-editor is part of the Planning.Domains initiative which aims at
-providing three pillars to the planning community: (1) an API for
-existing planning problems; (2) a planner-in-the-cloud service; and
-(3) an online PDDL editor. In contrast to \mypddl,
-editor.planning.domains does not support context-aware syntax
-highlighting: for example, editor.planning.domains highlights key
-words even if they are misplaced. In contrast, \mypddl is aware of the
-scope of a construct and highlights key words accordingly.
-
-The tool
-vscode-pddl\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
-offers a wide range of editing functions, such as syntax highlighting,
-code completion, and code snippets. Since it is implemented as a
-plugin for the editor Visual Studio Code, it can fall back on the the
-extensibility and maturity of this code editor. In contrast to
-\mypddl, it does not offer domain visualization nor an interface with
-a programming language in order to extend \pddl's limited modeling
-capacities.
-
-In sum, there is currently no tool available supporting all features
-of \pddl 3.1, nor all the steps in the modeling process.
+The online tool \epd allows for editing \pddl files in a web
+browser. Its feature comprise syntax highlighting, code folding,
+\pddl-specific auto-completion, and multi-tab support. The editor is
+part of the Planning.Domains initiative which aims at providing three
+pillars to the planning community: (1) an API to access existing \pddl
+domains and problems; (2) a planner-in-the-cloud service; and (3) an
+online PDDL editor. The online editor is also connected to the
+planner-in-the-cloud.
+
+\epd's syntax highlighting features are basic; so far, it does not
+support context-aware syntax highlighting: for example,
+editor.planning.domains highlights \pddl key words even if they are
+misplaced. In contrast, \mypddl is aware of the scope of a construct
+and highlights key words accordingly.
+
+The PDDL plugin
+vscode-\pddl\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
+for the editor Visual Studio Code offers a wide range of editing
+functions, such as syntax highlighting, code completion, and code
+snippets. It offers a mature planner integration and plan
+visualization. Thanks to a \pddl parser integration, it is possible to
+detect semantic errors immediately when they are made.
+
+\subsection{Critical Review}
+\label{sec-3-4}
+
+All the above-mentioned tools provide environments for the creation of
+\pddl code. Their advantages and disadvantages are reviewed in this
+section. At the end of each discussed feature, the approach that was
+used in \mypddl is introduced.
+
+\pddlstudio, \itsimple, and \epd for the most part do not build on
+existing editors and therefore cannot fall back on refined
+implementations of features, such as selection of tab size, defining
+custom key shortcuts, customizing the general look and feel, and
+bracket matching. In contrast, vscode-\pddl is integrated into the
+mature editor Visual Studio Code and can be used in combination with
+other plugins. To have both basic editor features and a high
+customizability, it was decided to use an existing, extensible text
+editor to integrate \mypddl into.
+
+The tools can also be compared in terms of their syntax highlighting
+capabilities. In \pddlmode for Emacs (up to \pddl\,2.2) and \epd (up
+to \pddl\,3.1), and vscode-\pddl (up to \pddl\,3.1) keywords,
+variables, and comments are highlighted. However, this is only done
+via pattern matching without controlling for context. This means that
+wherever the respective terms appear within the code they will get
+highlighted, regardless of the syntactical correctness. Therefore, it
+is useful when the knowledge engineer is familiar with \pddl syntax
+but can also be misleading if this is not the case. Different colors
+can be chosen by customizing Emacs and Visual Studio Code. \epd
+provides two fixed color schemes. \itsimple's syntax highlighting for
+\pddl 3.1 is, except for the \pddl version difference, equally as
+extensive as that of \pddlmode for Emacs but does not allow for any
+customization.  \pddlstudio has advanced syntax highlighting that
+distinguishes all different \pddl 1.2 constructs depending on the
+context and allows knowledge engineers to choose their preferred
+highlighting colors. One of the primary objectives of \mypddl is to
+help users in keeping track of their \pddl programs. As a means to
+this end, it was decided to also implement sophisticated,
+context-dependent syntax highlighting.
+
+Another feature that can be useful for fast programming is the ability
+to insert larger code skeletons or snippets. This allows the knowledge
+engineer to focus on the specific domain and problem characteristics
+instead of having to worry about the \pddl formalities. \pddlstudio
+does not support the insertion of code snippets. \itsimple features
+some code templates for predicates, derived predicates, functions,
+actions, constraints, types, comments, requirements, objects, and
+metrics. However, the templates are neither customizable nor
+extensible. \pddlmode for Emacs provides three larger skeletons: one
+for domains, one for problems, and one for actions. Further skeletons
+could be added. Both \epd and vscode-\pddl provide many code
+snippets. \mypddl aims to combine the best of these latter tools and
+support customizable and extensible snippets for domains, problems,
+types, predicates, functions, actions, and durative actions.
+
+\pddlstudio, \pddlmode for Emacs, and \epd do not provide
+visualization options. \itsimple, on the other hand, is based entirely
+on visually modeling domains and problems. Therefore, since the first
+version, the focus has mainly been on exporting from \textsc{uml.p} to
+\pddl. \mypddl is to reverse this design approach and enable type
+diagram visualization of some parts of the \pddl code. vscode-\pddl
+does not provide domain visualizationa but is able to visualize a
+found plan.
+
+Searching for errors can be one of the most time consuming parts of
+the design process. Hence, any tool that is able to help detect errors
+faster is of great value to the knowledge engineer. While \pddlmode
+for Emacs, \itsimple, and \epd facilitate error detection only by
+basic syntax highlighting, both \pddlstudio and vscode\-pddl are able
+to detect errors via a \pddl parser. In \mypddl, syntactic errors are
+not be highlighted by the syntax highlighting feature, while all
+correct \pddl code is to get highlighted.
+
+A major drawback of \pddlstudio and \pddlmode for Emacs especially is
+that they are not updated regularly to support the most recent \pddl
+versions. \pddlstudio's parser is only able to parse \pddl 1.2
+\cite{chomuttool}, while the latest \pddl version is 3.1. \pddl has
+significantly evolved since \pddl 1.2 and was extended in \pddl 2.1 to
+include \emph{durative actions} to model time dependent behaviors,
+\emph{numeric fluents} to model non-binary changes of the world state,
+and \emph{plan-metrics} to customize the evaluation of plans
+\cite{fox2003pddl2}. \pddlmode for Emacs is only compatible with \pddl
+versions up to 2.2, which introduced \emph{derived predicates} and
+\emph{timed initial predicates} \cite{edelkamp2004pddl2} but does not
+recognize later features like \emph{object-fluents}. It follows that
+the range of functions specified in the domain file cannot include
+object-types in addition to numbers. \itsimple, \epd, vscode-\pddl,
+and \mypddl support the latest \pddl version.
+
+An important features of any software is the possibility of extending
+and customizing it \cite{hui2003requirements}. \pddlstudio falls short
+of satisfying this requirement as the customization options are
+limited to the choice of font style and color of highlighted \pddl
+expressions. Furthermore, \pddlstudio is written as standalone
+program, meaning that there are no \pddl independent extensions. The
+same holds true for \itsimple. Since both Emacs and VS Code are
+established editors, \pddlmode and vscode-\pddl are highly and easily
+customizable and extensible. This is the other major reason why it was
+decided that \mypddl should be integrated into a existing, extensible,
+and customizable text editor. These requirements are intended to be
+met by Sublime Text, a text editor that offers such features as
+customizable key bindings, display of line numbers, and multi-line
+selection.
+
+All in all, \mypddl must be understood as complementary to the other
+existing knowledge engineering tools. \mypddl is distributed as a
+package for Sublime Text and provides context-aware syntax
+highlighting, code snippets, syntactic error detection, and type
+diagram visualization. Additionally, it allows for the automation of
+modeling tasks due to an interface with Clojure that supports the
+conversion of \textsc{pddl} code into Clojure code and vice
+versa. Therefore, \mypddl is intended to support both the initial
+design process of creating domains with code snippets, syntax
+highlighting and the Clojure interface and the later step of checking
+the validity of existing domains and problems with the type diagram
+generator. Lastly, the visualization capabilities of \mypddl are meant
+to facilitate collaboration among knowledge engineers.
+
+\begin{sidewaystable}[]
+\centering
+\scriptsize
+\caption[Comparison of knowledge engineeringtools]{\label{tool-comp}Comparison of knowledge engineering tools and their features.}
+\begin{tabularx}{\textwidth}{lX|llllll}
+  \tableheadline{Feature}             & \tableheadline{Function}                              & \tableheadline{\pddlstudio} & \tableheadline{\itsimple} & \tableheadline{\pddlmode} & \tableheadline{planning.domains} & \tableheadline{VS Code} & \tableheadline{\mypddl} \\
+  \hline
+  latest supp.
+\pddl version      & considering recent \pddl features                     & 1.2                         & 3.1                       & 2.2                           &   3.1    &  3.1    & 3.1       \\
+  syntax highlighting                 & supporting error detection and code navigation        & yes                         & basic                     & basic                         &   basic  & yes     & yes       \\
+  semantic error detection            & supporting error detection                            & yes                         & no                        & no                            &   no     & yes     & no        \\
+  automatic indentation               & supporting readability and navigation                 & no                          & no                        & yes                           &   no     & yes     & yes       \\
+  code completion                     & speeding-up the knowledge engineering process         & yes                         & no                        & yes                           &   yes    & yes     & yes       \\
+  code snippets                       & speeding-up the knowledge engineering process         & no                          & yes                       & yes                           &   basic  & yes     & yes       \\
+                                      & externalizing user's memory                           &                             &                           &                               &          &         &           \\
+  code folding                        & supporting keeping an overview of the code structure  & yes                         & no                        & yes                           &   yes    & yes     & yes       \\
+  domain visualization                & supporting fast understanding of the domain structure & no                          & no                        & no                            &   no     & no      & yes       \\
+  project management                  & supporting keeping an overview of associated files    & yes                         & yes                       & no                            &   yes    & yes     & yes       \\
+  \uml to \pddl translation           & supporting initial modeling                           & no                          & yes                       & no                            &   no     & no      & yes       \\
+  planner integration                 & allowing for convenient planner access                & basic                       & yes                       & no                            &   yes    & yes     & yes       \\
+  plan visualization                  & supporting understanding and crosschecking the plan   & no                          & yes                       & no                            &   no     & yes     & no        \\
+  dynamic analysis                    & supporting dynamic domain analysis                    & no                          & yes                       & no                            &   no     & no      & no        \\
+  declaration menu                    & supporting code navigation                            & no                          & no                        & yes                           &   no     & yes     & no        \\
+  interface with                      & automating tasks                                      & no                          & no                        & no                            &   no     & no      & yes       \\
+  programming language                & extending \pddl's modeling capabilities               &                             &                           &                               &          &         &           \\
+  customization features              & acknowledging individual needs and preferences        & basic                       & no                        & yes                           &  basic   & yes     & yes       \\
+\end{tabularx}
+\label{tab:comparison}
+\end{sidewaystable}
 
@@
 collaboration between knowledge engineers and the use of \pddl in
-real-world applications. \mypddl consists of the following integral
-parts:
+real-world applications. The modules of \mypddl are described in the
+next section.
 
@@
 \begin{description}
-\item[\mypddlide] is an integrated development environment (IDE) for
+  \setlength\itemsep{1em}
+\item[myPDDL-IDE] is an integrated development environment (IDE) for
   the use of \mypddl in the text and code editor \emph{Sublime
@@
   \mypddlsnippet and \textsc{-syntax} are devised explicitly for
-  \sublimetext, their integration is implicit. The other tools (-new,
-  -diagram, -distance) can be used independently of \sublimetext with
-  the command-line interface and any \pddl file, but were also
-  integrated into the editor.
-
-\item[\mypddlsyntax] is a context-aware syntax highlighting feature
-  for \sublimetext. It distinguishes all \pddl constructs up to
-  version 3.1. Using regular expressions that can recognize both the
-  start and the end of code blocks by means of a sophisticated pattern
-  matching heuristic, \mypddlsyntax identifies \pddl code blocks and
-  constructs and divides them into so called scopes, i.e. named
-  regions. \sublimetext colorizes the code elements via the assigned
-  scope names and in accordance with the current color scheme. These
-  scopes allow for a fragmentation of the \pddl files, so that
-  constructs are only highlighted if they appear in the correct
-  context. Thus missing brackets, misplaced expressions and misspelled
-  keywords are visually distinct and can be identified (see
-  Figure~\ref{fig:syntax}).
+  \sublimetext, their integration is implicit. The other tools
+  described below (\mypddl-new, -diagram, -distance, -plan) can be
+  used independently of \sublimetext via the command-line but can also
+  be called from the editor.
+
+\item[myPDDL-syntax] is a context-aware syntax highlighting feature
+  for \sublimetext. It recognizes all \pddl constructs up to version
+  3.1, such as comments, variables, names, and keywords and highlights
+  them in different colors. Using regular expressions and a
+  sophisticated pattern matching heuristic, it detects both the start
+  and the end of \pddl code blocks and constructs. It then divides
+  them into \emph{scopes}, that is, named regions. \sublimetext
+  colorizes the code elements via the assigned scope names and in
+  accordance with the current color scheme. These scopes allow for a
+  fragmentation of the \pddl files, so that constructs are only
+  highlighted if they appear in the correct context. Thus missing
+  brackets, misplaced expressions and misspelled keywords are visually
+  distinct and can be identified (see Figure~\ref{fig:syntax}).
 
@@
     \includegraphics[width=0.6\textwidth]{coffee-yes.pdf}
-    \caption{Syntax highlighting using \mypddlide. White
-      text contains errors.}
+    \caption{The figure shows the use of \mypddl in the text editor
+      Sublime Text. Syntax errors in the domain are detected by
+      \mypddlsyntax's context-aware syntax highlighting feature and
+      displayed in white.}
 \label{fig:syntax}
@@
 
-\item[\mypddlnew] helps to organize \pddl projects by generating the
-following folder structure:
-
-\begin{figure}[h] 
+\item[myPDDL-new] helps to organize \pddl projects. In many cases
+  \pddl domains are created ad hoc \cite{shah2013exploring}. However,
+  each implementation of a \pddl task specification comprises one
+  domain and at least one corresponding problem file. Since several
+  team members may be working on these files, keeping \pddl projects
+  organized will facilitate collaboration. An automatically created,
+  standardized project folder structure could facilitate the
+  collaboration between users and the maintenance of consistency
+  across projects. To this end, \mypddlnew creates the following
+  folder structure when creating a new \pddl project:
+  \begin{figure}[H]
   \dirtree{%
@@
 \end{figure}
-
-The domain file \texttt{domain.pddl} and the problem file
-\texttt{p01.pddl} initially contain corresponding \pddl skeletons
-which can also be customized. All problem files that are associated
-with one domain file are collected in the folder
-\texttt{problems/}. \texttt{README.md} is a Markdown file, which is
-intended for (but not limited to) information about the author(s) of
-the project, contact information, informal domain and problem
+  All of the templates to create the files can be customized and new
+  templates can be added. The domain file \texttt{domain.pddl} and the
+  problem file \texttt{p01.pddl} initially contain corresponding \pddl
+  skeletons. Additionally the project name is used as the domain name
+  within the files \texttt{domain.pddl} and \texttt{p01.pddl}. All
+  problem files that are associated with one domain file are collected
+  in the folder \texttt{problems/}. \texttt{README.md} is a Markdown
+  file, which is intended for information about the authors of the
+  project, contact information, informal domain and problem
   specifications, and licensing information.  Markdown files can be
-converted to \textsc{html} by various hosting services (like GitHub or
-Bitbucket). The basic planner integration \mypddlplan provided by the
-file \texttt{plan} is described below.
-
-\item[\mypddlsnippet] provides code skeletons, i.e. templates for
-  often used pddl constructs such as domains, problems, type and
+  converted to \textsc{html} by various hosting services (like GitHub
+  or Bitbucket). The basic planner integration \mypddlplan provided by
+  the file \texttt{plan} is described below.
+
+\item[myPDDL-snippet] provides code skeletons, that is, templates for
+  often used \pddl constructs such as domains, problems, type and
   function declarations, and actions. They can be inserted by typing a
-  triggering keyword. Almost all \pddl actions consist of these same parts. And this is only
-an example. Knowledge engineers have to use the same constructs again
-and again when writing and extending \pddl files. This is where code
-snippets can come in handy. To facilitate and speed up the implementation of
-standard constructs, \mypddl-snippet provides code skeletons, i.e.
-templates for often used \pddl constructs. They can be inserted by
-typing a triggering keyword (trigger). Table \ref{tab:snippets}
-displays descriptions of all available snippets and the corresponding
-trigger.
+  triggering keyword. Table \ref{tab:snippets} displays descriptions
+  of all available snippets and the corresponding trigger.
 
@@
   \centering
+  \caption[Available snippets in \mypddl-snippet]{\label{tab:snippets}The snippets that can be inserted into \pddl files by typing the trigger.}
 \begin{tabular}{ll}
@@
   action skeleton                     & \texttt{action}, \texttt{durative-action} \\
-\end{tabular}\caption[Available snippets in \mypddl-snippet]{\label{tab:snippets}The snippets that can be inserted into \pddl files by typing the trigger.}
-
+\end{tabular}
 \end{table}
 
-For example, typing \texttt{action} and pressing the tabulator key (this
-function defaults to the tabulator key, but the key can be customized)
+For example, typing \texttt{action} and pressing the tabulator key
 inserts the action skeleton in Listing \ref{ls:action-skeleton}. \pddl
@@
 
-Once the snippet has been inserted, skipping from blank to blank is
-enabled by pressing the tabulator key. This allows for a fast
-navigation within the snippet.
-
 Every snippet is stored in a separate file, located in the folder
 \texttt{Packages/PDDL/} of Sublime Text. New snippets can be added and
-existing snippets can be customized (by changing the template or the
-trigger) in this folder.
-
-\item[\mypddlclojure] provides a preprocessor for \pddl files to
+existing snippets can be customized by changing the templates in this
+folder.
+
+\item[myPDDL-clojure] provides a preprocessor for \pddl files to
   bypass \pddl's limited mathematical capabilities, thus reducing
@@
   is used to create more and more complex domains
-  \cite{goldman2012type,guerin2012academic}, needing the square root
-  function for a distance optimization problem or the logarithmic
-  function for modeling an engineering problem does not seem to be a
-  far-fetched scenario. However, \pddl's calculating capabilities are
-  limited \cite{parkinson2012increasing}. While these mathematical
-  operations are currently not supported by \pddl itself,
-  preprocessing \pddl files in a programming language and then
-  hardcoding the results back into the file seems to be a reasonable
-  workaround. With the help of such an interface, the modeling time
-  can reduced an even partly automated (see \ref{subsec:loc} the
-  distance calculator \mypddl-distance).  We decided to use Clojure
+  \cite{goldman2012type,guerin2012academic}, one might need the square
+  root function for a distance optimization problem or the logarithmic
+  function for modeling an engineering problem. However, \pddl's
+  calculating capabilities are limited
+  \cite{parkinson2012increasing}. While these mathematical operations
+  are currently not supported by \pddl itself, preprocessing \pddl
+  files in a programming language and then hardcoding the results back
+  into the file seem to be a reasonable workaround. With the help of
+  such an interface, the modeling time can be reduced.  We decided to
+  use the functional programming language Clojure
   \cite{hickey2008clojure}, a modern Lisp dialect that runs on the
@@
   facilitating input and output of the Lisp-style \pddl
-  constructs. The interface is provided as a Clojure library and based
-  on two methods:
+  constructs. Once a part is extracted and represented in Clojure, the
+  processing possibilities are diverse and the full capacities of
+  Clojure can be used. It can be used for generating \pddl constructs,
+  reading domain and problem files, handling, using and modifying the
+  input, and generating \pddl files as output.
+
+  The interface is provided as a Clojure library and based on two
+  methods described below.
 \begin{description}
-\item[{read-construct(keyword, file)}] Allows for the extraction of
-  code blocks from \pddl files. Listing \ref{src:read-construct} shows
-  an example in which the goal state of \emph{Gary's Huge Problem} is
-  extracted.
+\item[{read-construct(keyword, file)}] This methods allows for the extraction of
+  code blocks from \pddl files. The following code block shows an
+  example in which the goal state \texttt{(:goal (exploited
+    magicfailureapp))} is extracted.
 \end{description}
-\begin{description}
-\item[{add-construct(file, block, part)}] Provides a means for adding
-constructs to a specified code block in \pddl problem files. This
-is illustrated in Listing \ref{ls:using-interface} and
-\ref{ls:after-interface} where the predicate \texttt{(hungry gisela)} is
-added to the \texttt{(:init~...)} block of \emph{Gary's Huge Problem}.
-\end{description}
-
-\begin{listing}[H]
+Clojure command:
 \begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{clojure}
@@
 \end{minted}
-\caption[Extracting the goal state using the interface]{\label{src:read-construct}Extracting the goal state of \emph{Gary's Huge Problem} by using the interface.}
-\end{listing}
 
 \begin{description}
-\item[{add-construct(file, block, part)}] Provides a means for adding
-constructs to a specified code block in \pddl problem files. This
-is illustrated in Listing \ref{ls:using-interface} and
-\ref{ls:after-interface} where the predicate \texttt{(hungry gisela)} is
-added to the \texttt{(:init~...)} block of \emph{Gary's Huge Problem}.
+\item[{add-construct(file, block, part)}] This methods provides a
+  means for adding constructs to a specified code block in \pddl
+  problem files. This is illustrated in the following two code blocks
+  where the predicate \texttt{(hungry gisela)} is added to the
+  \texttt{(:init~...)} block.
 \end{description}
-
-\begin{listing}[H]
+Clojure command:
 \begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{clojure}
@@
 \end{minted}
-\caption[Adding a predicate using the
-interface]{\label{ls:using-interface}Adding the predicate
-  \texttt{(hungry~ gisela)} to the \texttt{(:init~...)} block of
-  \emph{Gary's Huge Problem} using the PDDL/Clojure interface.}
-\end{listing}
-
-\begin{listing}[H]
+Updated \pddl file:
 \begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
@@
 \end{minted}
-\caption[Modified problem file]{\label{ls:after-interface}The modified
-  part of the problem file to which the predicate \texttt{(hungry gisela)} has
-  been added.}
-\end{listing}
-
-\item[\mypddldistance] provides special preprocessing functions for
-  distance calculations. Let us assume that Gary from Chapter \ref{ch:basics} wants the
-\emph{Hacker World} to be more realistic. Every object shall now have
-a location specified by a x and y coordinate. Imagine this domain now
-including a further precondition in the action \texttt{eat-pizza}, so
-that Gary can only eat the pizza if it is at most an arm's length
-away. While the location of objects can be implemented using the
-predicate \texttt{(location ?o - object ?x ?y - number)}, with
-\texttt{x} and \texttt{y} being the spatial coordinates of an object,
-calculating the Euclidean distance requires using the square root
-function ($\sqrt{}$). Unfortunately, \pddl 3.1 supports only four
-arithmetic operators (\texttt{+}, \texttt{-}, \texttt{/}, \texttt{*}).
-How then is it possible to prevent hackers from starving?
-
-\cite{parkinson2012increasing} describe a workaround for this
-drawback. By writing an action \texttt{calculate-sqrt}, they bypass
-the missing square root function by making use of the Babylonian root
-method. Although this method approximates the square root function, it
-requires many iterations and would most likely have an adverse effect
-on plan generation \cite{parkinson2012increasing}.
+
+\item[myPDDL-distance] provides special preprocessing functions for
+  distance calculations. In some domains, every object needs a
+  location specified by $x$ and $y$ coordinates.  While the location
+  of objects can be implemented using the predicate \texttt{(location
+    ?o - object ?x ?y - number)}, with \texttt{x} and \texttt{y} being
+  the spatial coordinates of an object, calculating the Euclidean
+  distance requires using the square root function. However, \pddl 3.1
+  supports only the four basic arithmetic operators.
+
+  Parkinson and Longstaff~\cite{parkinson2012increasing} describe a
+  workaround for this drawback. By writing an action
+  \texttt{calculate-sqrt}, they bypass the missing square root
+  function by making use of the Babylonian root method. Although this
+  method approximates the square root function, it requires many
+  iterations and would most likely have an adverse effect on plan
+  generation \cite{parkinson2012increasing}.
 
@@
 interface between \pddl and Clojure as a distance calculator,
-implemented in the tool \mypddl-distance. It reads a problem file into
+implemented in the tool \mypddldistance. It reads a problem file into
 Clojure and extracts all locations, defined in the \texttt{(:init
@@
 of the problem file, using the predicate
-\texttt{(distance~?o1~?o2~-~object ?n~-~number)}\footnote{This
-  predicate has to be declared in the \texttt{(:predicates ...)} block
-  of the corresponding domain file.}, which specifies the distance
-between two objects. Listing \ref{ls:before-mypddl-distance} and
-\ref{ls:after-mypddl-distance} show the \texttt{(:init ...)} code
+\texttt{(distance~?o1~?o2~-~object ?n~-~number)}, which specifies the
+distance between two objects. Listing \ref{ls:before-mypddl-distance}
+and \ref{ls:after-mypddl-distance} show the \texttt{(:init ...)} code
 block of a extended version of \emph{Gary's Huge Problem} before and
@@
 The calculator works on any arity of the specified location predicate,
-so that locations can be specified one, two and three dimensionally
-and even used in higher dimensions.
-
-However, this preprocessing of problem file also has a drawback, apart
-from the time required to calculate possibly unused distances. If the
-number of locations is \(n\), the number of calculated distances is
-\(n^2\), because every location has a distance to every other location
-(and to itself). The calculated distances have to be stored in the
-\pddl problem file, potentially requiring a lot of space. Therefore, a
-sensible next step would be to extend \pddl by increasing its
-mathematical expressivity \cite{parkinson2012increasing}, perhaps by
-declaring a requirement \texttt{:math} that specifies further
-mathematical operations.
-
-For domains with spatial components, the distance of objects is often
-important and should not be omitted in the domain model. However,
-calculating distances from coordinates requires the square root
-function, which is not supported by \pddl (it only supports the four
-basic arithmetic operators). More sophisticated calculations can be
-achieved with the supported operators, but the solutions are rather
-inefficent and inelegant \cite{parkinson2012increasing}. By
-calculating the distances offline and including them as additional
-predicates in the problem file using \mypddldistance, the distances
-between objects are given to the planner as part of the problem
-description.
+so that locations can be specified in 1D, 2D, 3D, and even used in
+higher dimensions.
+
+A disadvantage of this method is that the calculated distances have to
+be stored in the \pddl problem file, potentially requiring many lines
+of code. If the number of locations is \(n\), the number of calculated
+distances is \(n^2\), so that every location has a distance to every
+other location and itself. Therefore, a sensible next step would be
+to extend \pddl by increasing its mathematical expressivity
+\cite{parkinson2012increasing}, perhaps by declaring a requirement
+\texttt{:math} that specifies further mathematical operations.
 
@@
 
-\item[\mypddldiagram] generates a \textsc{png} image from a \pddl
+\item[myPDDL-diagram] generates a \textsc{png} image from a \pddl
   domain file as shown in Figure~\ref{fig:diagram}. The diagrammatic
@@
   able to simplify the communication and collaboration between
-  developers. In the process of generating the diagrams, an optional
-  copy of the \pddl file can be created, providing a simple version
-  control system. In the diagram, types are represented with
-  boxes, with every box consisting of two parts:
+  developers. In the diagram, types are represented with boxes, with
+  every box consisting of two parts:
 
@@
 In order to create the diagram, \mypddl-diagram utilizes dot from the
-Graphviz package\footnote{Graphviz is a collection of programs for
-  drawing graphs. This includes dot, a scriptable, graphing tool, that
-  is able to generate hierarchical drawings of directed graphs in a
-  variety of output formats (e.g. \textsc{png}, \textsc{pdf},
-  \textsc{svg}). The input to dot are text files, written in the
-  \textsc{dot} language.} \cite{ellson2002graphviz} and takes the
-following steps:
+Graphviz package~\cite{ellson2002graphviz} and takes the following
+steps:
 
@@
 
-Every time \mypddl-diagram is invoked, these steps are executed and
-the names of the saved files are extended by an ascending revision
-number. Thus, one cannot only identify associated \pddl, \textsc{dot}
-and \textsc{png} files, but also use this feature for basic revision
-control. Figure~\ref{fig:mypddl-new-project-folder} displays the
-folder structure after invoking \mypddl-diagram twice on the
-\emph{Hacker World}.
+Every time \mypddl-diagram is invoked, these steps are executed and,
+optionally, the names of the saved files are extended by an ascending
+revision number. Thus, one cannot only identify associated \pddl,
+\textsc{dot} and \textsc{png} files, but also use this feature for
+basic revision control. Figure~\ref{fig:mypddl-new-project-folder}
+displays the folder structure after invoking \mypddl-diagram twice on
+the \emph{Hacker World}.
 
@@
     \includegraphics[width=1\textwidth]{dot-diagram}
-    \caption{Type diagram generated by \mypddldiagram}
+    \caption{The type diagram generated by \mypddldiagram helps to
+      grasp the relationship between types in the domain
+      file. Additionally, it displays all predicates that use the
+      corresponding type at least once as a parameter.}
     \label{fig:diagram}
@@
 
-\item[\mypddlplan] is a basic planner integration for \mypddl. After
+\item[myPDDL-plan] is a basic planner integration for \mypddl. After
   creating a new project with \mypddlnew, the file \texttt{plan} in
@@
   can be specified in the file \texttt{plan} or by editing the
-  template. Due to the versatility of shell scripts, any planner can
-  be used and arbitrary command line options can be specified.
-
+  templates of \mypddlnew. Due to the versatility of shell scripts,
+  any planner can be used and arbitrary command line options can be
+  specified. The planner can be invoked from Sublime Text or via the
+  command line.
 \end{description}
 
-\subsection{Design Principles}
-
-As guidelines for design decisions, we used the seven criteria for
-knowledge engineering tools proposed by Shah et
-al. \cite{shah2013knowledge} as well as general usability
-principles. Operationality instantiates, whether the generated models
-can improve the planning performance. This is not a design principle
-for \mypddl, because we assume that \mypddl does not improve the
-quality (with respect to planning performance) of the resulting \pddl
-specifications. Therefore, we replaced this criterion with functional
-suitability from the \textsc{iso/iec} 25010 standard, which is defined
-as “the degree to which the software product provides an appropriate
-set of functions for specified tasks and user objectives”
-(\textsc{iso} 25010 6.1.1). \mypddl supports the current version 3.1
-of \pddl. It encompasses and exceeds most of the functionality of the
-existing tools. It specifically supports basic editor features with a
-high customizability as well as visualization support.
-
-\begin{description}
-
-\item[Collaboration:] With the growing importance of team work and
-  team members not necessarily working in the same building, or in the
-  same country, there is an increasing need for tools supporting the
-  collaboration effort. In developing \mypddl, this need was sought to
-  be met by \mypddldiagram. Complex type hierarchies can be hard to
-  overlook, especially if they were constructed by someone
-  else. Therefore, a good way of tackling this problem seemed to be by
-  providing a means to visualize such hierarchies in the form of type
-  diagrams.
-
-\item[Experience:] \mypddl was designed specifically for users with a
-  background in AI, but not necessarily in \pddl. The tools are
-  similar to standard software engineering tools and should thus be
-  easily learnable. The user evaluation (Section TODO) confirms that
-  \mypddl helps novices in \pddl to master planning task modeling. In
-  addition, it is also possible to customize \mypddl so as to adapt
-  its look and feel to other programs one is already familiar with, or
-  simply to make it more enjoyable to use. The project
-  site\footnote{\url{http://pold87.github.io/myPDDL}} provides \mypddl
-  video introductions and a manual to get started quickly.
-
-\item[Efficiency:] All \mypddl tools are intended to increase the
-  efficiency with which \pddl files are created. \mypddlsnippet
-  enables the fast creation of large and correct code skeletons that
-  only need to be complemented. \mypddlsyntax can reduce the time
-  spent on searching errors. Code folding allows users to hide
-  currently irrelevant parts of the code and automatic indentation
-  increases its readability. To easily keep track of all the parts of
-  a project, folders are automatically created and named with
-  \mypddlnew. \mypddlclojure and \textsc{-distance} allow
-  for a straightforward inclusion of numerical values in the problem
-  definition.
-
-\item[Debugging:] \mypddl\textsc{-syntax} highlights all syntactically
-  correct constructs and leaves all syntactical errors
-  non-highlighted. In contrast, \pddlmode for Emacs and itSimple only
-  provide basic syntax highlighting for emphasizing the
-  structure. \pddlstudio explicitly detects errors, but the user is
-  immediately prompted when an error is detected. Often, such error
-  messages are premature, for example, just because the closing
-  parenthesis was not typed yet, does not mean it was
-  forgotten. \mypddl indicates errors in a more subtle way: syntactic
-  errors are simply not highlighted, while all correct \pddl code
-  is. The colors are customizable, so that users can choose how
-  prominent the highlighting sticks out.
-
-\item[Maintenance:] The possibility to maintain \pddl files is a key
-  aspect of \mypddl. The automatically generated type diagram
-  (\mypddldiagram) gives an overview of the domain structure and
-  thereby serves as a continuous means of documentation. Helping to
-  understand foreign code, though, it follows logically that
-  \mypddldiagram also helps in coming back and changing one’s own
-  models if some time has elapsed since they were last edited. The
-  basic revision control feature of \mypddldiagram keeps track of
-  changes, making it easy to revert to a previous domain
-  version. Furthermore, \mypddlnew encourages adhering to an organized
-  project structure and stores corresponding files at the same
-  location.  The automatically created readme file can induce the user
-  to provide further information and documentation about the \pddl
-  project.
-
-\item[Support:] \mypddlide can be installed using \sublimetext's
-  Package
-  Control\footnote{\url{https://sublime.wbond.net/about}}. This allows
-  for an easy installation and staying up-to-date with future
-  versions. In order to provide global access and with it the
-  possibility for developing an active community, the project source
-  code is hosted on
-  GitHub\footnote{https://github.com/Pold87/myPDDL}. Additionally, the
-  project site provides room for discussing features and reporting
-  bugs.
-
-\end{description}
+In order to provide easy installation and maintenance, \mypddlide can
+be installed using \sublimetext's Package
+Control\footnote{\url{https://sublime.wbond.net/about}}.  The project
+source code is hosted on
+GitHub\footnote{\url{https://github.com/Pold87/myPDDL}}, providing the
+possibility to actively participate in the design
+process. \mypddlclojure is hosted at
+\footnote{\url{https://github.com/Pold87/pddl-clojure-interface}}
+Additionally, the project
+site\footnote{\url{http://pold87.github.io/myPDDL/}} provides room for
+discussing features and reporting bugs.
 
@@
 
-To assess the utility of \mypddl, we used the criteria listed in
-Section TODO. The functional suitability was evaluated using a
-benchmark validation, comparing \mypddl's functionality with the tools
-described in Section 2. The criteria collaboration, experience,
-efficiency, and debugging were evaluated in a user test. We analyzed
-the user performance both with and without using \mypddlsyntax and
-\mypddldiagram. The \mypddl components supporting maintenance are the
-same ones that are used in the user test, but their long-term usage is
-difficult to evaluate. The support criterion depends primarily on the
-infrastructure, which has been established as explained in 3.2.
-
-\subsection{Benchmark Validation}
-
-Functional suitability encompasses the set of functions to meet the
-user objectives. The tools of Section 2 basically all follow the same
-objectives as \mypddl: creating \pddl domains and problems. The
-features offered by each tool are summarized in Table
-\ref{tool-comp}. Besides supporting the latest \pddl version, a
-strength of \mypddl is its high customizability, which comes with the
-\sublimetext editor. Being the only one of the six tools capable of
-visualizing parts of the \pddl code, it must be understood as
-complementary to \itsimple, which takes the opposite approach of
-transforming \uml diagrams into \pddl files. The fact that \mypddl
-does not check for semantic errors is not actually a drawback because
-planners usually detect semantic errors. All in all, \mypddl strives
-to support the planning task engineer during all phases of the
-modeling process. Additionally, it features some unique tools, such as
-domain visualization and an interface with a programming language. It
-can therefore be concluded that \mypddl provides an appropriate set of
-functions for developing \pddl files and is thus functionally
-suitable.
-
-\begin{sidewaystable}[]
-\centering
-\footnotesize
-\begin{tabularx}{\textwidth}{lX|llll}
-  \tableheadline{Feature}             & \tableheadline{Function}                              & \tableheadline{\pddlstudio} & \tableheadline{\itsimple} & \tableheadline{\pddlmode} & \tableheadline{editor.planning.domains} & \tableheadline{PDDL - Visual Studio} & \tableheadline{\mypddl} \\
-  \hline
-  latest supported \pddl version      & considering recent \pddl features                     & 1.2                         & 3.1                       & 2.2                           &   3.1    &  3.1    & 3.1       \\
-  syntax highlighting                 & supporting error detection and code navigation        & yes                         & basic                     & basic                         &   basic  & yes     & yes       \\
-  semantic error detection            & supporting error detection                            & yes                         & no                        & no                            &   no     & yes     & no        \\
-  automatic indentation               & supporting readability and navigation                 & no                          & no                        & yes                           &   no     & yes     & yes       \\
-  code completion                     & speeding-up the knowledge engineering process         & yes                         & no                        & yes                           &   yes    & yes     & yes       \\
-  code snippets                       & speeding-up the knowledge engineering process         & no                          & yes                       & yes                           &   basic  & yes     & yes       \\
-                                      & externalizing user's memory                           &                             &                           &                               &          &         &           \\
-  code folding                        & supporting keeping an overview of the code structure  & yes                         & no                        & yes                           &   yes    & yes     & yes       \\
-  domain visualization                & supporting fast understanding of the domain structure & no                          & no                        & no                            &   no     & no      & yes       \\
-  project management                  & supporting keeping an overview of associated files    & yes                         & yes                       & no                            &   yes    & yes     & yes       \\
-  \uml to \pddl code translation      & supporting initial modeling                           & no                          & yes                       & no                            &   no     & no      & yes       \\
-  planner integration                 & allowing for convenient planner access                & basic                       & yes                       & no                            &   yes    & yes     & yes       \\
-  plan visualization                  & supporting understanding and crosschecking the plan   & no                          & yes                       & no                            &   no     & yes     & no        \\
-  dynamic analysis                    & supporting dynamic domain analysis                    & no                          & yes                       & no                            &   no     & no      & no        \\
-  declaration menu                    & supporting code navigation                            & no                          & no                        & yes                           &   no     & yes     & no        \\
-  interface with programming language & automating tasks                                      & no                          & no                        & no                            &   no     & no      & yes       \\
-                                      & extending \pddl's modeling capabilities               &                             &                           &                               &          &      &              \\
-  customization features              & acknowledging individual needs and preferences        & basic                       & no                        & yes                           &  basic   & yes     & yes       \\
-\end{tabularx}\caption[Comparison of knowledge engineering
-tools]{\label{tool-comp}Comparison of knowledge engineering tools and their features.}
-
-\end{sidewaystable}
+To assess the utility of \mypddl, we evaluated its performance in
+terms of collaboration, experience, efficiency, and debugging in a
+user test. We analyzed the user performance both with and without
+using \mypddlsyntax and \mypddldiagram.
 
@@
 
-No earlier than 24 hours before the experiment was to take place,
-participants received the web
-link\footnote{\url{https://www.youtube.com/watch?v=Uck-K8VnNOU&list=PL3CZzLUZuiIMWEfJxy-G6OxYVzUrvjwuV} (tutorial is in German)}
-to a 30-minute interactive video tutorial on AI planning and
-\pddl. This method was chosen in order not to pressure the participant
-with the presence of an experimenter when trying to understand the
-material.
+24 hours before the experiment was to take place, participants
+received the web
+link\footnote{Tutorial in German: \url{https://www.youtube.com/watch?v=Uck-K8VnNOU&list=PL3CZzLUZuiIMWEfJxy-G6OxYVzUrvjwuV}} to a 30-minute interactive video tutorial
+on AI planning and \pddl. This method was chosen in order not to
+pressure the participant with the presence of an experimenter when
+trying to understand the material.
 
@@
 
-We designed \mypddl to support knowledge engineers in all steps of
-developing a planning domain. Additionally, it support engineers to
-understand, modify, and extend existing \pddl planning domains.
-
-This was realized with a set of tools comprising code editing
-features, namely syntax highlighting and code snippets, a type diagram
-generator, and a distance calculator. To also have all tools
-accessible from one place, they were made available in the Sublime
-Text editor. The different needs and requirements of knowledge
-engineers are met by the modular, extensible and customizable
-architecture of the toolkit and Sublime Text. The evaluation of
-\mypddl has shown some initial evidence that it allows a faster
-understanding of the domain structure, which could be beneficial for
-the maintenance and application of existing task specifications and
-for the communication between engineers. Users perceive it as easy and
-enjoyable to use, and the increase in their performance when using
-\mypddl underpins their subjective impressions. It is hoped that the
-planning community not only accepts but actually uses \mypddl
-intensely, leading to extensive feedback and continuous improvements.
+We designed \mypddl to support knowledge engineers in creating,
+understanding, modifying, and extending planning domains. \mypddl's
+code editing features such as syntax highlighting and code snippets,
+as well as a type diagram generator and an interface with the
+programming language Clojure, can help in the various stages of
+working with planning domains. \mypddl's extensible and customizable
+architecture helps to fulfill the different preferences and
+requirements of knowledge engineers. In the conducted user test,
+\mypddl user were able to grasp the domain structure of a \pddl file
+more quickly than non-users and also found more errors in a
+. Moreover, the users found it as easy and pleasant to use.
 
@@
 directions. The interface between \pddl and Clojure offers a basis
-for creating dynamic planning scenarios. Applications could be the modeling of
-learning and forgetting (by adding facts to or retracting facts from a
-\pddl file) or the modeling of an ever changing real world via dynamic
-predicate lists. Another way of putting the interface to use would be
-by making the planning process more interactive, allowing for the
-online interception of planning software in order to account for the
-needs and wishes of the end user.
+for creating dynamic planning scenarios. Applications could be the
+modeling of learning and forgetting by adding facts to or retracting
+facts from a \pddl file or the modeling of an ever changing real world
+via dynamic predicate lists. Another way of putting the interface to
+use would be by making the planning process more interactive, allowing
+for the online interception of planning software in order to account
+for the needs and wishes of the end user. Since \mypddl is command
+line-based, interfaces with other editors could be developed. There is
+a basic integration into the code editor
+Atom\footnote{\url{https://github.com/Pold87/myPDDL-Atom}}.
 

### 2019-07-03 paper.tex
upd

@@
   interface between \pddl and Clojure as a distance calculator,
-implemented in the tool \mypddldistance. It reads a problem file into
-Clojure and extracts all locations, defined in the \texttt{(:init
-  ...)} code block. The Euclidean distances between these locations
-are then calculated and written back into a new and now extended copy
-of the problem file, using the predicate
-\texttt{(distance~?o1~?o2~-~object ?n~-~number)}, which specifies the
-distance between two objects. Listing \ref{ls:before-mypddl-distance}
-and \ref{ls:after-mypddl-distance} show the \texttt{(:init ...)} code
-block of a extended version of \emph{Gary's Huge Problem} before and
-after using \mypddl-distance, respectively.
-
-The calculator works on any arity of the specified location predicate,
-so that locations can be specified in 1D, 2D, 3D, and even used in
-higher dimensions.
-
-A disadvantage of this method is that the calculated distances have to
-be stored in the \pddl problem file, potentially requiring many lines
-of code. If the number of locations is \(n\), the number of calculated
-distances is \(n^2\), so that every location has a distance to every
-other location and itself. Therefore, a sensible next step would be
-to extend \pddl by increasing its mathematical expressivity
-\cite{parkinson2012increasing}, perhaps by declaring a requirement
-\texttt{:math} that specifies further mathematical operations.
-
-\begin{listing}[H]
+  implemented in the tool \mypddldistance. It reads a problem file
+  into Clojure and extracts all locations, defined in the
+  \texttt{(:init ...)} code block. The Euclidean distances between
+  these locations are then calculated and written back into a new and
+  now extended copy of the problem file, using the predicate
+  \texttt{(distance~?o1~?o2~-~object ?n~-~number)}, which specifies
+  the distance between two objects. The code blocks below show the
+  \texttt{(:init ...)} block of a \pddl problem file before and after using
+  \mypddl-distance.
 \begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
@@
 \end{minted}
-\caption[Excerpt of the extended problem file before using
-\mypddl-distance]{\label{ls:before-mypddl-distance}Excerpt of the
-  \texttt{(:init~...)} block of the extended file \emph{Gary's Huge
-    Problem} before using \mypddl-distance.}
-\end{listing}
-
-\begin{listing}[H]
 \begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
@@
 \end{minted}
-\caption[Excerpt of the extended problem file after using
-\mypddl-distance]{\label{ls:after-mypddl-distance}After the
-  application of \mypddl-distance, the calculated distances are
-  inserted in the \texttt{(:init ...)} code block in a copy of the
-  problem file.}
-\end{listing}
+
+The calculator works on any arity of the specified location predicate,
+so that locations can be specified in 1D, 2D, 3D, and even used in
+higher dimensions.
+
+A disadvantage of this method is that the calculated distances have to
+be stored in the \pddl problem file, potentially requiring many lines
+of code. If the number of locations is \(n\), the number of calculated
+distances is \(n^2\), so that every location has a distance to every
+other location and itself. Therefore, a sensible next step would be
+to extend \pddl by increasing its mathematical expressivity
+\cite{parkinson2012increasing}, perhaps by declaring a requirement
+\texttt{:math} that specifies further mathematical operations.
 
@@
     bottom of the bars show the percentage of users that completed the
-    task successfully. The question can be found in the Appendix
-    (\ref{splisus}, \ref{store}).}\label{fig:task-completions-agg}
+    task successfully. The questions can be found in the
+    Appendix~\ref{splisus} and
+    \ref{store}.}\label{fig:task-completions-agg}
 \end{figure}
@@
 
+\appendix
+
+\section{Tasks}
+\label{sub:tasks}
+\subsection{Deliberately erroneous Logistics Domain}
+\label{logistics}
+\begin{minted}[fontsize=\tiny,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
+;;;; Logistics domain
+
+(define (domain ?logistics)
+
+  (:requirements
+    :types)
+
+  (:typing truck airplane motorboat - vehicle
+           package vehicle suitcase furniture - thing
+           airport garage station - location
+           car1 car 2 car3 - vehicle
+           city location thing - object)
+
+  (:predicates (in-city ?l - location ?c - city)
+               (at ?obj - thing ?l - location)
+               (key ?v - vehicle) = true
+               (full ?v - vehicle)
+               (in ?p - package ??veh - vehicle))
+
+  (:action drive
+    :parameters (?t - truck ?from ?to - location ?c - city)
+    :precondition (and (at ?tr ?from)
+                       (in-city ?from ?c)
+                       (incity ?to ?c))
+    :effect (and (not (at ?t ?from))
+                 (at ?t ?to)))
+
+  (:action fly
+    :parameters (?a - airplane ?from ?to - airport)
+    :precondition (at ?a ?from)
+    :effect (and (n0t (at ?a ?from))
+                 (at ?a ?to)))
+
+  (:action fuel
+    :parameters (?v - vehicle ?c - city ?to airport)
+    :precondition (and (not (full ?v))
+                       (in-city ?to ?c)
+                       (at ?v ?to))
+    :effect (full ?v))
+
+  (:action load
+    parameters: (?v - vehicle ?p - package ?l - location)
+    precondition: (and (?v ?l)
+                       (at ?p ?l))
+    :effect (and (ay ?p ?l)
+                 (in ?p ?v)))
+
+  (:action unload
+    :parameters (?v - vehicle p - package ?l - location)
+    :precondition (and (at ?v ?l)
+                           ?p ?v)
+    :effects (and (not (in ?p ?v))
+                  (at ?p - ?l))))
+                \end{minted}
+The original file can be downloaded at
+\url{http://ipc.informatik.uni-freiburg.de/PddlExtension}:
+
+\subsection{Deliberately erroneous Coffee Domain}
+\label{coffee}
+\begin{minted}[fontsize=\scriptsize,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
+(define COFFEE
+
+  (requirements
+    :typing)
+
+  (:types room - location
+                 robot human _ agent
+                 furniture door - (at ?l - location)
+                 kettle ?coffee cup water - movable
+                 location agent movable - object)
+
+  (:predicates (at ?l - location ??o - object)
+               (have ?m - movable ?a - agent)
+               (hot ?m - movable) = true
+               (on ?f - furniture ?m - movable))
+
+  (:action boil
+    :parameters (?m - movable \$k - kettle ?a - agent)
+    :preconditions (have ?m ?a)
+    :effect (hot ?m))
+
+  (:action grip-some
+    :parameters (?m - movable ?r - robot ?f - _furniture ?l - location)
+    :precondition (and (at ?l ?r)
+                       (on ?fu ?m)
+                       (at ?l ?f))
+    :effect (and (have ?m ?r)))
+
+  (:action move
+    :parameters: (?m - movable ?a - agent ?from ?to - location)
+    :precondition (or (änd (at ?from ?a)
+                           (at ?from ?m))
+                           (and (at ?from ?m)
+                                (location ?from ?a)))
+    :effect (and (not (at ?from ?m))
+                 (at ?to ?m)))
+
+  (:action change-room
+    :parameters (?from-r ?to-r - room ?a - agent)
+    :precondition (at ?fromr ?a) 
+    :effect (and (not (at ?from-r ?a))
+                 (at ?tor ?a)))
+
+  (:action prep-coffee
+    :parameters (?a - agent ?c - cjp ?w - water ?cof - coffee)
+    :precondition (and (have ?c ?a)
+                       (hot ?w))
+    :effect (have ?cof ?a))
+
+  (:action ?hand-over
+    :parameters (?m - movable ?a1 - agent ?a2 - agent)
+    :precondition (have ?m ?a1))
+    :effect (and (not (have ?m ?a1))
+                 (have ?m ?a2))))
+\end{minted}
+
+\subsection{Planet Splisus}
+\label{splisus}
+\begin{minted}[fontsize=\scriptsize,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
+(define (domain splisus)
+
+  (:requirements :typing)
+
+  (:types splis - gid
+          spleus - splos
+          schprok schlok - splus
+          rud mekle - lech
+          hulpf hurpf - hupf
+          sipsi flipsi hupf - splis
+          schmok schkok - splus
+          gid splos splus - ruffisplisus
+          merle - hupf
+          ruffisplisus mak lech - object)
+
+  (:predicates (father-of ?r1 - ruffisplisus ?r2 - ruffisplisus)
+               (married ?s1 - splos ?s2 - splis)
+               (has-weapon ?h - sipsi)
+               (dead ?r1 - ruffisplisus)
+               (at ?l - lech ?r - ruffisplisus))
+
+  (:action kill
+    :parameters (?l - lech ?r1 - ruffisplisus ?s - splis)
+    :precondition (and (at ?l ?r1)
+                       (at ?l ?s)
+                       (married ?r1 ?s)
+                       (has-weapon ?s))
+    :effect (and (dead ?r1)
+            (not (married ?r1 ?s)))))
+\end{minted}
+
+Please answer the following five questions on the society and
+structure of Planet Splisus:
+\begin{enumerate}
+\item Are all \emph{Flipsis} also of the type \emph{Ruffisplisus}?
+\item Are all \emph{Merles} also \emph{Splus}?
+\item Can a \emph{Spleus} be married to a \emph{Schlok}?
+\item Only theoretically: Could a \emph{Hurpf} murder a \emph{Spleus}?
+\item Let us assume there are three categories of object types on
+  Splisus: places, beings and food. Match the three object types
+  \emph{Ruffisplisus}, \emph{Mak} and \emph{Lech} with these
+  categories.
+\end{enumerate}
+
+\subsection{Store}
+\label{store}
+\begin{minted}[fontsize=\scriptsize,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
+(define (domain store)
+
+  (:requirements :typing)
+
+  (:types lala lila - zahls
+          blisis blusis - ultri
+          iltre lula - nulls
+          zahls schwinds - knozi
+          minis - lala
+          ultri sopple schmitzl - lila
+          ultres raglos wexis - lola
+          kosta - nulls
+          nulls spax - minis
+          lola - zahls
+          knozi schmus - object)
+
+  (:predicates (product ?k - knozi) ; Produkt
+               (workplace ?l1 - lola ?l2 - lala) 
+               (product-at ?l1 - lola ?l2 - lila) 
+               (cashier ?k - knozi) 
+               (customer ?s - spax)
+               (owns ?l - lila ?s - spax))
+
+  (:action sell
+    :parameters (?p - lila ?z - zahls ?l - lola ?w - wexis ?s - spax)
+    :precondition (and (product ?p)
+                       (cashier ?z)
+                       (product-at ?l ?p)
+                       (customer ?s))
+    :effect (and (product-at ?w ?p)
+                 (not (product-at ?l ?p))
+                 (owns ?p ?s))))
+\end{minted}
+Please answer the following five questions concerning the environment
+store:
+\begin{enumerate}
+\item Are objects of the type \emph{Lula} also of the type
+    \emph{Minis}?
+\item Are \emph{Spax} and \emph{Schmus} \emph{Zahls}?
+\item Is it possible for an Iltre to work at a
+    \emph{workplace} of the type \emph{Knozi}?
+\item Only theoretically: Could a \emph{Lala} sell a
+    \emph{Schmitzl} to a \emph{Kosta}?
+\item Let us assume our domain \emph{store} models a
+    grocery store. There are three categories: humans, products, and
+    places. Can you match these world terms with the object types
+    \emph{lila}, \emph{lala}, and \emph{lola} from the domain?
+\end{enumerate}
+
 \end{document}

### 2019-07-19 paper.tex
Sent to Mauro version

@@
 \usepackage{booktabs}
+
 % More space between rows:		
@@
 \definecolor{dhscodebg}{rgb}{0.95,0.95,0.95} % listing background
-\DeclareUnicodeCharacter{2212}{-}
+%\DeclareUnicodeCharacter{2212}{-}
 
@@
 \newcommand{\pddlmode}{\textsc{pddl}-mode\xspace}
+\newcommand{\vscode}{vscode-\textsc{pddl}\xspace}
 \newcommand{\pddl}{\textsc{pddl}\xspace}
@@
 goal \cite{helmert2008understanding}. It can be both a tool to create
-automated systems and a means to support and understand human behavior
-\cite{konar1999artificial}. However, the effectiveness of planning
-largely depends on the quality of the problem formalization
-\cite{shah2013knowledge,keps2014}. To ensure a standardized modeling
-format, \pddl (\emph{Planning Domain Definition Language})
+automated systems and a means to support and understand human
+behavior~\cite{konar1999artificial}. However, the effectiveness of
+planning largely depends on the quality of the problem formalization
+\cite{shah2013knowledge}. To ensure a standardized modeling format,
+\pddl (\emph{Planning Domain Definition Language})
 \cite{mcdermott1998pddl} was developed and has become the de facto
@@
 numerous other possibilities and could also be used to further
-automate the modeling process. All of the features were integrated
-into the customizable and extensible Sublime Text \cite{sublimetext3}
-editor. Since the main aim in the development of the toolkit was for
-it to be easy to use and maintain, it is evaluated with regard to
-these criteria. The usability was assessed by means of a user test
-with eight subjects that had no prior experience with AI planning. The
-results indicate that both error-detection and the understanding of a
-given domain can be facilitated by \mypddl. This chapter is an
-extended version of work published in a previous paper~\cite{StrKir2014:aai}.
+automate the modeling process. A basic planner integration allows for
+quickly running a desired planner. All of the features were integrated
+into the customizable and extensible Sublime
+Text\footnote{\url{http://www.sublimetext.com/}} editor. Since the
+main aim in the development of the toolkit was for it to be easy to
+use and maintain, it is evaluated with regard to these criteria. The
+usability was assessed by means of a user test with eight subjects
+that had no prior experience with AI planning. The results indicate
+that both error-detection and the understanding of a given domain can
+be facilitated by \mypddl.
 
@@
   child { node {\textbf{-new}\\Create new \pddl project}}
+  child { node {\textbf{-snippet}\\Code templates}}
   child { node {\textbf{-syntax}\\Context-aware syntax highlighting}}
-    child { node {\textbf{-snippet}\\Code templates}}
   child { node {\textbf{-clojure}\\Preprocessor for \pddl files via Clojure}}
@@
 
-The remainder of this chapter is structured as
-follows. Section~\ref{sec:related-work} compares \pddl knowledge
-engineering tools to lay a foundation for
+This chapter is an extended version of work published in a previous
+paper~\cite{StrKir2014:aai}. The remainder of this chapter is
+structured as follows. Section~\ref{sec:related-work} compares \pddl
+knowledge engineering tools to lay a foundation for
 \mypddl. Section~\ref{sec:mypddl} describes the different modules of
@@
 
-\pddlstudio \cite{chomuttool,plch2012inspect} is an IDE
-(\emph{integrated development environment}) for creating and managing
-\textsc{pddl} projects, that is, a collection of \textsc{pddl}
-files. Its main features are syntax highlighting, error detection,
-context sensitive code completion, code folding, project management,
-and planner integration. Many of these features are based on a parser,
-which continuously analyzes the code and divides it into syntactic
+\pddlstudio \cite{plch2012inspect} is an IDE (\emph{integrated
+  development environment}) for creating and managing \textsc{pddl}
+projects, that is, a collection of \textsc{pddl} files. Its main
+features are syntax highlighting, error detection, context sensitive
+code completion, code folding, project management, and planner
+integration. Many of these features are based on a parser, which
+continuously analyzes the code and divides it into syntactic
 elements. These elements and the way in which they relate to each
@@
 type of predicate parameters, misspelled predicates, etc.)
-errors. This means that \pddlstudio can detect errors due to a
+errors. This means that \pddlstudio can detect errors based on a
 mismatch between domain and problem file in real time. The code
@@
 code blocks. In this case only the first line of the block is
-displayed. All these above mentioned features of \pddlstudio utilize
-the parser. To keep track of all files, to display them in a tree
-structure, and to save them upon compilation, a project manager is
-integrated. Furthermore, this feature is necessary to detect interfile
-errors and to use the code completion functionality. Lastly, a
-command-line interface allows the integration of planners in order to
-run and compare different planning software.
+displayed. Lastly, a command-line interface allows the integration of
+planners in order to run and compare different planning software.
 
@@
   Language}). In the process leading up to \itsimple,
-\textsc{uml.p}---\textsc{uml} in a Planning Approach---was proposed, a
-\textsc{uml} variant specifically designed for modeling planning
-domains and problems \cite{vaquero2006use}.
-
-The main purpose of \itsimple is supporting knowledge
-engineers in the initial stages of the design phase by making tools
-available that help transform the informality of the real world to
-formal specifications of domain models. The professed aim of the
-project is to provide a means to a \enquote{disciplined process of
-  elicitation, organization and analysis of requirements}
-\cite{vaquero2005itsimple}. However, subsequent design stages are also
-supported. Once domain and problem models have been created,
-\textsc{pddl} representations can be generated from the \textsc{uml.p}
-diagrams, edited, and then used as input to a number of different
-integrated planning systems. Compared to \pddlstudio, \itsimple has a
-different approach to planner integration: domain and problem can be
-fed to the planner by pressing a button, whereas in \pddlstudio, the
-user has to know and input commands in a command-line interface.
+\textsc{uml.p}---\textsc{uml} in a Planning Approach---was proposed,
+which is a \textsc{uml} variant specifically designed for modeling
+planning domains and problems \cite{vaquero2006use}.
+
+The main purpose of \itsimple is supporting knowledge engineers in the
+initial stages of the design phase by making tools available that help
+transform the informality of the real world to formal specifications
+of domain models. The professed aim of the project is to provide a
+means to a \enquote{disciplined process of elicitation, organization
+  and analysis of requirements} \cite{vaquero2005itsimple}. However,
+subsequent design stages are also supported. Once domain and problem
+models have been created, \textsc{pddl} representations can be
+generated from the \textsc{uml.p} diagrams, edited, and then used as
+input to a number of different integrated planning systems.
 
@@
 not affect the \uml model and \uml models have to be modeled manually,
-meaning that they cannot by generated using \pddl. Starting in version
-4.0 \cite{vaquero2012itsimple4} \itsimple expanded its features to
-allow the creation of \pddl projects from scratch (i.e. without the
-\uml to \pddl translation process). Thus far, the \pddl editing
-features are basic. A minimal syntax highlighting feature recognizes
-\pddl keywords, variables, and comments. Furthermore, \itsimple
-provides templates for \pddl constructs, such as requirement
-specifications, predicates, actions, initial, and goal definitions.
+meaning that they cannot by generated using \pddl.
 
@@
 \pddl-specific auto-completion, and multi-tab support. The editor is
-part of the Planning.Domains initiative which aims at providing three
-pillars to the planning community: (1) an API to access existing \pddl
-domains and problems; (2) a planner-in-the-cloud service; and (3) an
-online PDDL editor. The online editor is also connected to the
-planner-in-the-cloud.
-
-\epd's syntax highlighting features are basic; so far, it does not
-support context-aware syntax highlighting: for example,
-editor.planning.domains highlights \pddl key words even if they are
-misplaced. In contrast, \mypddl is aware of the scope of a construct
-and highlights key words accordingly.
-
-The PDDL plugin
-vscode-\pddl\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
-for the editor Visual Studio Code offers a wide range of editing
-functions, such as syntax highlighting, code completion, and code
-snippets. It offers a mature planner integration and plan
-visualization. Thanks to a \pddl parser integration, it is possible to
-detect semantic errors immediately when they are made.
+part of the Planning-Domains\footnote{\url{http://planning.domains/}}
+initiative which aims at providing three pillars to the planning
+community: (1) an API to access existing \pddl domains and problems;
+(2) a planner-in-the-cloud service which be be accessed via a RESTful
+API; and (3) an online \pddl editor. The online editor is also
+connected to the planner in the cloud.
+
+The \pddl plugin
+\vscode\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
+for the editor VS Code (\emph{Visual Studio Code}) offers a wide range
+of editing functions, such as syntax highlighting, code completion,
+code folding, and code snippets. It offers a mature planner
+integration and plan visualization. Thanks to a \pddl parser
+integration, it is possible to detect semantic errors immediately when
+they are made.
 
@@
 custom key shortcuts, customizing the general look and feel, and
-bracket matching. In contrast, vscode-\pddl is integrated into the
+bracket matching. In contrast, \vscode is integrated into the
 mature editor Visual Studio Code and can be used in combination with
@@
 capabilities. In \pddlmode for Emacs (up to \pddl\,2.2) and \epd (up
-to \pddl\,3.1), and vscode-\pddl (up to \pddl\,3.1) keywords,
-variables, and comments are highlighted. However, this is only done
-via pattern matching without controlling for context. This means that
-wherever the respective terms appear within the code they will get
-highlighted, regardless of the syntactical correctness. Therefore, it
-is useful when the knowledge engineer is familiar with \pddl syntax
-but can also be misleading if this is not the case. Different colors
-can be chosen by customizing Emacs and Visual Studio Code. \epd
-provides two fixed color schemes. \itsimple's syntax highlighting for
-\pddl 3.1 is, except for the \pddl version difference, equally as
-extensive as that of \pddlmode for Emacs but does not allow for any
-customization.  \pddlstudio has advanced syntax highlighting that
-distinguishes all different \pddl 1.2 constructs depending on the
-context and allows knowledge engineers to choose their preferred
-highlighting colors. One of the primary objectives of \mypddl is to
-help users in keeping track of their \pddl programs. As a means to
-this end, it was decided to also implement sophisticated,
-context-dependent syntax highlighting.
-
-Another feature that can be useful for fast programming is the ability
-to insert larger code skeletons or snippets. This allows the knowledge
-engineer to focus on the specific domain and problem characteristics
-instead of having to worry about the \pddl formalities. \pddlstudio
-does not support the insertion of code snippets. \itsimple features
-some code templates for predicates, derived predicates, functions,
-actions, constraints, types, comments, requirements, objects, and
+to \pddl\,3.1), and \vscode (up to \pddl\,3.1) keywords, variables,
+and comments are highlighted. However, this is only done via pattern
+matching without controlling for context. This means that wherever the
+respective terms appear within the code they will get highlighted,
+regardless of the syntactical correctness. Different colors can be
+chosen by customizing Emacs and Visual Studio Code. \epd provides two
+fixed color schemes. \itsimple's syntax highlighting for \pddl 3.1 is,
+except for the \pddl version difference, equally as extensive as that
+of \pddlmode for Emacs but does not allow for any customization.
+\pddlstudio has advanced syntax highlighting that distinguishes all
+different \pddl 1.2 constructs depending on the context and allows
+knowledge engineers to choose their preferred highlighting colors. One
+of the primary objectives of \mypddl is to help users in keeping track
+of their \pddl programs. As a means to this end, it was decided to
+also implement sophisticated, context-dependent syntax highlighting.
+
+Another feature that can be useful for fast development is the ability
+to insert larger code skeletons or snippets. \pddlstudio does not
+support the insertion of code snippets. \itsimple features some code
+templates for predicates, derived predicates, functions, actions,
+constraints, types, comments, requirements, objects, and
 metrics. However, the templates are neither customizable nor
@@
 for domains, one for problems, and one for actions. Further skeletons
-could be added. Both \epd and vscode-\pddl provide many code
+could be added. Both \epd and \vscode provide many code
 snippets. \mypddl aims to combine the best of these latter tools and
@@
 version, the focus has mainly been on exporting from \textsc{uml.p} to
-\pddl. \mypddl is to reverse this design approach and enable type
-diagram visualization of some parts of the \pddl code. vscode-\pddl
-does not provide domain visualizationa but is able to visualize a
-found plan.
+\pddl and to visualize plans. \mypddl is to reverse this design
+approach and enable type diagram visualization of some parts of the
+\pddl code. \vscode does not provide domain visualization but is able
+to visualize a found plan.
 
@@
 for Emacs, \itsimple, and \epd facilitate error detection only by
-basic syntax highlighting, both \pddlstudio and vscode\-pddl are able
-to detect errors via a \pddl parser. In \mypddl, syntactic errors are
-not be highlighted by the syntax highlighting feature, while all
-correct \pddl code is to get highlighted.
+basic syntax highlighting, both \pddlstudio and \vscode are able to
+detect errors via a \pddl parser. In \mypddl, a different approach is
+taken and syntactic errors are \emph{not} highlighted by the syntax
+highlighting feature, while all correct \pddl code \emph{is}
+highlighted.
 
@@
 that they are not updated regularly to support the most recent \pddl
-versions. \pddlstudio's parser is only able to parse \pddl 1.2
-\cite{chomuttool}, while the latest \pddl version is 3.1. \pddl has
-significantly evolved since \pddl 1.2 and was extended in \pddl 2.1 to
-include \emph{durative actions} to model time dependent behaviors,
-\emph{numeric fluents} to model non-binary changes of the world state,
-and \emph{plan-metrics} to customize the evaluation of plans
-\cite{fox2003pddl2}. \pddlmode for Emacs is only compatible with \pddl
-versions up to 2.2, which introduced \emph{derived predicates} and
-\emph{timed initial predicates} \cite{edelkamp2004pddl2} but does not
-recognize later features like \emph{object-fluents}. It follows that
-the range of functions specified in the domain file cannot include
-object-types in addition to numbers. \itsimple, \epd, vscode-\pddl,
-and \mypddl support the latest \pddl version.
-
-An important features of any software is the possibility of extending
-and customizing it \cite{hui2003requirements}. \pddlstudio falls short
-of satisfying this requirement as the customization options are
+versions. \pddlstudio's parser is only able to parse \pddl 1.2, while
+the latest \pddl version is 3.1. \pddl has significantly evolved since
+\pddl 1.2 and was extended in \pddl 2.1 to include \emph{durative
+  actions} to model time dependent behaviors, \emph{numeric fluents}
+to model non-binary changes of the world state, and
+\emph{plan-metrics} to customize the evaluation of plans. \pddlmode
+for Emacs is only compatible with \pddl versions up to 2.2, which
+introduced \emph{derived predicates} and \emph{timed initial
+  predicates} but does not recognize later features like
+\emph{object-fluents}. It follows that the range of functions
+specified in the domain file cannot include object-types in addition
+to numbers. \itsimple, \epd, \vscode, and \mypddl support the latest
+\pddl version.
+
+\pddlstudio falls short of customization options since they are
 limited to the choice of font style and color of highlighted \pddl
 expressions. Furthermore, \pddlstudio is written as standalone
-program, meaning that there are no \pddl independent extensions. The
+program, meaning that there are no \pddl-independent extensions. The
 same holds true for \itsimple. Since both Emacs and VS Code are
-established editors, \pddlmode and vscode-\pddl are highly and easily
-customizable and extensible. This is the other major reason why it was
-decided that \mypddl should be integrated into a existing, extensible,
-and customizable text editor. These requirements are intended to be
-met by Sublime Text, a text editor that offers such features as
-customizable key bindings, display of line numbers, and multi-line
-selection.
+established editors, \pddlmode and \vscode are highly customizable and
+extensible. This is the other major reason why it was decided that
+\mypddl should be integrated into a existing, extensible, and
+customizable text editor. These requirements are met by Sublime Text,
+a text editor that offers a wide variety of features and plugins.
 
@@
 \begin{tabularx}{\textwidth}{lX|llllll}
-  \tableheadline{Feature}             & \tableheadline{Function}                              & \tableheadline{\pddlstudio} & \tableheadline{\itsimple} & \tableheadline{\pddlmode} & \tableheadline{planning.domains} & \tableheadline{VS Code} & \tableheadline{\mypddl} \\
+  \tableheadline{Feature}             & \tableheadline{Function}                              & \tableheadline{\pddlstudio} & \tableheadline{\itsimple} & \tableheadline{\pddlmode} & \tableheadline{planning.domains} & \tableheadline{\vscode} & \tableheadline{\mypddl} \\
   \hline
@@
   project management                  & supporting keeping an overview of associated files    & yes                         & yes                       & no                            &   yes    & yes     & yes       \\
-  \uml to \pddl translation           & supporting initial modeling                           & no                          & yes                       & no                            &   no     & no      & yes       \\
-  planner integration                 & allowing for convenient planner access                & basic                       & yes                       & no                            &   yes    & yes     & yes       \\
+  \uml to \pddl translation           & supporting initial modeling                           & no                          & yes                       & no                            &   no     & no      & no       \\
+  planner integration                 & allowing for convenient planner access                & yes                       & yes                       & no                            &   yes    & yes     & yes       \\
   plan visualization                  & supporting understanding and crosschecking the plan   & no                          & yes                       & no                            &   no     & yes     & no        \\
@@
     Text}\footnote{\url{http://www.sublimetext.com}}. Since
-  \mypddlsnippet and \textsc{-syntax} are devised explicitly for
-  \sublimetext, their integration is implicit. The other tools
-  described below (\mypddl-new, -diagram, -distance, -plan) can be
-  used independently of \sublimetext via the command-line but can also
-  be called from the editor.
-
-\item[myPDDL-syntax] is a context-aware syntax highlighting feature
-  for \sublimetext. It recognizes all \pddl constructs up to version
-  3.1, such as comments, variables, names, and keywords and highlights
-  them in different colors. Using regular expressions and a
-  sophisticated pattern matching heuristic, it detects both the start
-  and the end of \pddl code blocks and constructs. It then divides
-  them into \emph{scopes}, that is, named regions. \sublimetext
-  colorizes the code elements via the assigned scope names and in
-  accordance with the current color scheme. These scopes allow for a
-  fragmentation of the \pddl files, so that constructs are only
-  highlighted if they appear in the correct context. Thus missing
-  brackets, misplaced expressions and misspelled keywords are visually
-  distinct and can be identified (see Figure~\ref{fig:syntax}).
-
-  \begin{figure}
-    \centering
-    \includegraphics[width=0.6\textwidth]{coffee-yes.pdf}
-    \caption{The figure shows the use of \mypddl in the text editor
-      Sublime Text. Syntax errors in the domain are detected by
-      \mypddlsyntax's context-aware syntax highlighting feature and
-      displayed in white.}
-\label{fig:syntax}
-  \end{figure}
+  \mypddlsnippet and \textsc{-syntax} (see below) are devised
+  explicitly for \sublimetext, their integration is implicit. The
+  other tools described below (\mypddl-new, -diagram, -distance,
+  -plan) can be used independently of \sublimetext via the
+  command-line but can also be called from the editor.
 
@@
 specifications, and licensing information.  Markdown files can be
-  converted to \textsc{html} by various hosting services (like GitHub
-  or Bitbucket). The basic planner integration \mypddlplan provided by
-  the file \texttt{plan} is described below.
+converted to \textsc{html} by various hosting services like GitHub or
+Bitbucket. The basic planner integration \mypddlplan provided by the
+file \texttt{plan} is described below.
 
@@
 For example, typing \texttt{action} and pressing the tabulator key
-inserts the action skeleton in Listing \ref{ls:action-skeleton}. \pddl
-constructs with a specified arity can be generated by adding the arity
-number to the trigger (\texttt{p2} would insert the binary predicate
-template \texttt{(pred-name~?x~-~object~?y~-~object)}.
-
-Every snippet is stored in a separate file, located in the folder
-\texttt{Packages/PDDL/} of Sublime Text. New snippets can be added and
-existing snippets can be customized by changing the templates in this
-folder.
+inserts a skeleton to specify an action. \pddl constructs with a
+specified arity can be generated by adding the arity number to the
+trigger (\texttt{p2} would insert the binary predicate template
+\texttt{(pred-name~?x~-~object~?y~-~object)}.
+
+Every snippet is stored in a separate file, located in the packages
+folder of Sublime Text. New snippets can be added and existing
+snippets can be customized by changing the templates in this folder.
+
+\item[myPDDL-syntax] is a context-aware syntax highlighting feature
+  for \sublimetext. It recognizes all \pddl constructs up to version
+  3.1, such as comments, variables, names, and keywords and highlights
+  them in different colors. Using regular expressions and a
+  sophisticated pattern matching heuristic, it detects both the start
+  and the end of \pddl code blocks and constructs. It then divides
+  them into \emph{scopes}, that is, named regions. \sublimetext
+  colorizes the code elements via the assigned scope names and in
+  accordance with the current color scheme. These scopes allow for a
+  fragmentation of the \pddl files, so that constructs are only
+  highlighted if they appear in the correct context. Thus missing
+  brackets, misplaced expressions and misspelled keywords are visually
+  distinct and can be identified (Figure~\ref{fig:syntax}).
+
+  \begin{figure}
+    \centering
+    \includegraphics[width=0.6\textwidth]{coffee-yes.pdf}
+    \caption{The figure shows the use of \mypddl in the text editor
+      Sublime Text. Syntax errors in the domain are detected by
+      \mypddlsyntax's context-aware syntax highlighting feature and
+      displayed in white.}
+\label{fig:syntax}
+  \end{figure}
 
@@
   root function for a distance optimization problem or the logarithmic
-  function for modeling an engineering problem. However, \pddl's
-  calculating capabilities are limited
-  \cite{parkinson2012increasing}. While these mathematical operations
-  are currently not supported by \pddl itself, preprocessing \pddl
-  files in a programming language and then hardcoding the results back
-  into the file seem to be a reasonable workaround. With the help of
-  such an interface, the modeling time can be reduced.  We decided to
-  use the functional programming language Clojure
-  \cite{hickey2008clojure}, a modern Lisp dialect that runs on the
-  Java Virtual Machine (\textsc{jvm}) \cite{lindholm2014java},
+  function for modeling an engineering problem. While these
+  mathematical operations are currently not supported by \pddl itself,
+  preprocessing \pddl files in a programming language and then
+  hardcoding the results back into the file seem to be a reasonable
+  workaround. With the help of such an interface, the modeling time
+  can be reduced.  We decided to use the functional programming
+  language Clojure \cite{hickey2008clojure}, a modern Lisp dialect,
   facilitating input and output of the Lisp-style \pddl
@@
   processing possibilities are diverse and the full capacities of
-  Clojure can be used. It can be used for generating \pddl constructs,
+  Clojure are available. It can be used for generating \pddl constructs,
   reading domain and problem files, handling, using and modifying the
@@
 \begin{description}
-\item[{read-construct(keyword, file)}] This methods allows for the extraction of
-  code blocks from \pddl files. The following code block shows an
-  example in which the goal state \texttt{(:goal (exploited
-    magicfailureapp))} is extracted.
+\item[{read-construct(keyword, file)}] This methods allows for the
+  extraction of code blocks from \pddl files. The following code block
+  shows an example in which the goal state \texttt{(:goal (exploited
+    magicfailureapp))} is extracted from a \pddl problem file.
 \end{description}
@@
   means for adding constructs to a specified code block in \pddl
-  problem files. This is illustrated in the following two code blocks
-  where the predicate \texttt{(hungry gisela)} is added to the
-  \texttt{(:init~...)} block.
+  domain and problem files. This is illustrated in the following two
+  code blocks where the predicate \texttt{(hungry gisela)} is added to
+  the \texttt{(:init~...)} block.
 \end{description}
@@
   \mypddl-distance.
+
+  Before:
 \begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
@@
      \end{minted}
+     After:
 \begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{text}
@@
 
-\item[myPDDL-diagram] generates a \textsc{png} image from a \pddl
-  domain file as shown in Figure~\ref{fig:diagram}. The diagrammatic
-  representation of textual information helps to quickly understand
-  the connection of hierarchically structured items and should thus be
-  able to simplify the communication and collaboration between
-  developers. In the diagram, types are represented with boxes, with
-  every box consisting of two parts:
+\item[myPDDL-diagram] generates a \textsc{png} image based on the type
+  hierarchy of a \pddl domain file (Figure~\ref{fig:diagram}). The
+  diagrammatic representation of textual information helps to quickly
+  understand the connection of hierarchically structured items and
+  should thus be able to simplify the communication and collaboration
+  between developers. In the diagram, types are represented with
+  boxes, with every box consisting of two parts:
 
   \begin{itemize}
+    \setlength\itemsep{0.5em}
   \item The header displays the name of the type.
@@
 Generalization relationships express that every subtype is also an
-instance of the illustrated super type (e.g. ``a laptop \emph{is a}
-computer''). This relationship is indicated in the diagram with an
-arrow from the subtype (here: \emph{laptop}) to the super type (here:
-\emph{computer}).
+instance of the illustrated super type (e.g. ``a hacker \emph{is a}
+person). This relationship is indicated in the diagram with an
+arrow from the subtype (here: \emph{hacker}) to the super type (here:
+\emph{person}).
 
@@
 \begin{enumerate}
+  \setlength\itemsep{0.5em}
 \item A copy of the domain file is stored in the folder
@@
 \textsc{dot} and \textsc{png} files, but also use this feature for
-basic revision control. Figure~\ref{fig:mypddl-new-project-folder}
-displays the folder structure after invoking \mypddl-diagram twice on
-the \emph{Hacker World}.
+basic revision control.
 
@@
 possibility to actively participate in the design
-process. \mypddlclojure is hosted at
-\footnote{\url{https://github.com/Pold87/pddl-clojure-interface}}
-Additionally, the project
+process. Additionally, \mypddlclojure is hosted on GitHub
+\footnote{\url{https://github.com/Pold87/pddl-clojure-interface}} as
+well as a standalone version to call the functions from a the command
+line\footnote{\url{https://github.com/Pold87/pddl-clojure-interface-standalone}}.
+The \mypddl project
 site\footnote{\url{http://pold87.github.io/myPDDL/}} provides room for
@@
 0.6). All participants were required to have basic experience with at
-least one Lisp dialect (in order not to be confused with the many
-parentheses), but no experience with \pddl or AI planning in general.
+least one Lisp dialect in order not to be confused with the many
+parentheses, but no experience with \pddl or AI planning in general.
 
@@
 diagram generator. A within subjects design was considered most suited
-dueto the small number of participants. Therefore, it was necessary to
-construct two tasks (matched in difficulty) for each of these two
+due to the small number of participants. Therefore, it was necessary to
+construct two tasks matched in difficulty for each of these two
 types to compare the effects of having the tools available. Each
@@
   the domains, all of which could be facilitated with the type diagram
-  generator. One of the five questions (Question 4) also required
-  looking into the code. Participants were told that they should not
-  feel pressured to answer quickly, but to not waste time either. Also
-  they were asked to say their answer out loud as soon as it became
-  evident to them. They were not told that the time it took them to
-  come up with an answer was recorded, since this could have made them
-  feel pressured and thus led to more false answers.
+  generator. One of the five questions (Question 4, see
+  Appendix~\ref{splisus} and \ref{store}) also required looking into
+  the code. Participants were told that they should not feel pressured
+  to answer quickly, but to not waste time either. Also they were
+  asked to say their answer out loud as soon as it became evident to
+  them. They were not told that the time it took them to come up with
+  an answer was recorded, since this could have made them feel
+  pressured and thus led to more false answers.
 
@@
   and six layers in depth, 20 and 21 types). The domains were also
-  matched in length and overall complexity (five and six predicates
+  matched in length and overall complexity: five and six predicates
   with approximately the same distribution of arities, one action with
   four predicates in the precondition and two and three predicates in
-  the effect).
+  the effect.
 
@@
 \begin{itemize}
-\item Debugging Tasks
-\nopagebreak
-  To compare differences in the debugging tasks, a paired sample
-  $t$-test was used; normality was tested with a Shapiro-Wilk
-    test. to compare the arithmetic means ($M$s) of detected
-  errors. The test was performed two-tailed, since syntax highlighting
-  might both help or hinder the participants. Arithmetic standard
-  deviations ($SD$s) were calculated for each condition.
-
-\item Type Hierarchy Tasks
+\item Debugging Tasks: \nopagebreak To compare differences in the
+  debugging tasks, a paired sample $t$-test was used; normality was
+  tested with a Shapiro-Wilk test. To compare the arithmetic means
+  ($M$s) of detected errors, the test was performed two-tailed, since
+  syntax highlighting might both help or hinder the
+  participants. Arithmetic standard deviations ($SD$s) were calculated
+  for each condition.
+
+\item Type Hierarchy Tasks:
 \nopagebreak
@@
     \caption{Overview of geometric means ($GM$s), geometric standard deviations ($GSD$s), degrees of freedom ($df$), $t$-values, and $p$-values.
-Please note, that the calculation for Q4 is based on only two paired data values ($df = 1$).
-Please note further that this table only considers paired data values, this means only if a participant answered the question correctly in both domains, the data value is considered (since \emph{paired} $t$-tests are calculated).
+The calculation for Q4 is based on only two paired data values ($df = 1$).
+This table only considers paired data values, this means only if a participant answered the question correctly in both domains, the data value is considered (since \emph{paired} $t$-tests are calculated).
 In contrast, Figure \ref{fig:task-completions-agg} displays the geometric means for all correct answers.}
@@
   Question 4 can probably be attributed to the complexity of this
-  question (see Appendix, \ref{splisus} and \ref{store}): in contrast
+  question (see Appendix~\ref{splisus} and \ref{store}): in contrast
   to the other four questions, here, participants were required to
@@
 code editing features such as syntax highlighting and code snippets,
-as well as a type diagram generator and an interface with the
-programming language Clojure, can help in the various stages of
-working with planning domains. \mypddl's extensible and customizable
-architecture helps to fulfill the different preferences and
-requirements of knowledge engineers. In the conducted user test,
-\mypddl user were able to grasp the domain structure of a \pddl file
+as well as a type diagram generator, an interface with the programming
+language Clojure, and a planner integration can help in the various
+stages of working with \pddl domains. \mypddl's extensible and
+customizable architecture helps to fulfill the different preferences
+and requirements of knowledge engineers. In the conducted user test,
+\mypddl users were able to grasp the domain structure of a \pddl file
 more quickly than non-users and also found more errors in a
-. Moreover, the users found it as easy and pleasant to use.
+deliberately erroneous domain file. Moreover, the users found the
+tools easy and pleasant to use.
 
 In future work, \mypddl's set of features could be extended in several
-directions. The interface between \pddl and Clojure offers a basis
-for creating dynamic planning scenarios. Applications could be the
+directions. The interface between \pddl and Clojure offers a basis for
+creating dynamic planning scenarios. Applications could be the
 modeling of learning and forgetting by adding facts to or retracting
@@
 for the online interception of planning software in order to account
-for the needs and wishes of the end user. Since \mypddl is command
-line-based, interfaces with other editors could be developed. There is
-a basic integration into the code editor
+for the needs and wishes of the end user. Since many features of
+\mypddl can be called via the command line, interfaces with other
+editors could be developed. So far, there is a basic integration with
+the code editor
 Atom\footnote{\url{https://github.com/Pold87/myPDDL-Atom}}.
@@
 \bibliography{mybibliography}
-
+\newpage
 \appendix

### 2020-08-14 paper.tex
final version

@@
 \usepackage{rotating}
-\usepackage{caption}
 \usepackage{booktabs}
@@
 \definecolor{dhscodebg}{rgb}{0.95,0.95,0.95} % listing background
-%\DeclareUnicodeCharacter{2212}{-}
+\DeclareUnicodeCharacter{2212}{-}
 
@@
 
-Being a key aspect of artificial intelligence, \emph{planning} is
+Being a key aspect of artificial intelligence (AI), \emph{planning} is
 concerned with devising a sequence of actions to achieve a desired
-goal \cite{helmert2008understanding}. It can be both a tool to create
-automated systems and a means to support and understand human
-behavior~\cite{konar1999artificial}. However, the effectiveness of
-planning largely depends on the quality of the problem formalization
-\cite{shah2013knowledge}. To ensure a standardized modeling format,
-\pddl (\emph{Planning Domain Definition Language})
-\cite{mcdermott1998pddl} was developed and has become the de facto
-standard for the description of planning tasks
-\cite{ilghami2005extension}. The discipline that deals with the
-integration of world information into a computer system via a human
-expert is called knowledge engineering \cite{feigenbaum1983fifth}.
-While automated planning could save vast amounts of time if everything
-works as intended, creating the planning task specifications is a
-complex task that can be error-prone and cumbersome.
-
-In this chapter, we describe \mypddl
-(Figure~\ref{fig:mypddl-overview}), a knowledge engineering toolkit
-that supports the knowledge engineer in the entire design cycle of
-specifying planning tasks. In the initial stages, it allows for the
-creation of structured \pddl projects that should encourage a
-disciplined design process. With the help of snippets, that is, code
-templates, often used syntactic constructs can be inserted into \pddl
-files. A syntax highlighting feature that speeds up the
-error-detection can come in handy in intermediate
-stages. Understanding the textual representation of complex type
-hierarchies in domain files can be confusing, so an additional tool
-enables their visualization. \pddl's limited modeling capabilities
-were bypassed by developing an interface that converts \pddl code into
-code of the functional programming language Clojure
+goal~\cite{helmert2008understanding}. AI planning has made remarkable
+progress in solving planning problems in large state spaces that would
+be impossible for humans to handle. The International Planning
+Competition\footnote{\url{http://www.icaps-conference.org/index.php/Main/Competitions/}}
+has led to a number of open source planners that are ready to be used
+by practitioners and researchers outside the AI planning field.
+
+However, the effectiveness of planning largely depends on the quality
+of the problem formalization~\cite{shah2013knowledge}. \pddl
+(\emph{Planning Domain Definition Language})~\cite{mcdermott1998pddl}
+is the de facto standard for the description of planning
+tasks~\cite{ilghami2005extension}. It divides the description of a
+planning task into a domain model and problem descriptions: the
+description of a household with its objects and locations would be a
+domain, with possible tasks such as making breakfast, cleaning the
+windows, or changing a light bulb. Someone with a non-planning
+background, for example, from robotics, has to get used to the \pddl
+syntax and its possibilities to describe the world. She also has to
+keep track of the facts in the domain model and the different planning
+tasks. While automated planning can save vast amounts of time to find
+a valid solution, creating the planning task specifications is a
+complex, error-prone, and cumbersome task. An ill-defined problem is
+often the reason for finding suboptimal plans or no plan at all.  The
+household domain in this example comes with another challenge. For
+many tasks the distances between objects or other numerical input may
+be necessary. \pddl is by its very nature as a planning language
+designed for symbolic specifications. To use numerical data
+efficiently, it must often be preprocessed.
+
+In this chapter, we describe
+\mypddl~(Figure~\ref{fig:mypddl-overview}), a knowledge engineering
+toolkit that supports knowledge engineers in the entire design cycle
+of specifying planning tasks without having to become an expert in AI
+planning. In the initial stages, it allows for the creation of
+structured \pddl projects that should encourage a disciplined design
+process. With the help of snippets, that is, code templates, often
+used syntactic constructs can be inserted into \pddl files. A syntax
+highlighting feature that speeds up the error-detection supports
+intermediate stages. Understanding the textual representation of
+complex type hierarchies in domain files can be confusing, so an
+additional tool enables their visualization. \pddl's limited modeling
+capabilities were bypassed by developing an interface that converts
+\pddl code into code of the functional programming language Clojure
 \cite{hickey2008clojure} and vice versa.  Within this project, the
@@
 into the customizable and extensible Sublime
-Text\footnote{\url{http://www.sublimetext.com/}} editor. Since the
-main aim in the development of the toolkit was for it to be easy to
-use and maintain, it is evaluated with regard to these criteria. The
-usability was assessed by means of a user test with eight subjects
-that had no prior experience with AI planning. The results indicate
-that both error-detection and the understanding of a given domain can
-be facilitated by \mypddl.
+Text\footnote{\url{http://www.sublimetext.com/}} editor.
+
+Since the main aim in the development of the toolkit was for it to be
+easy to use and maintain, it is evaluated with regard to these
+criteria. Another aim was to make planning more accessible in
+real-life tasks and to enable inexperienced users to get started with
+planning problems. Therefore, \mypddl's usability was assessed by
+means of a user test with eight subjects that had no prior experience
+with AI planning. The results show that \mypddl facilitates both
+error-detection and the understanding of a given domain.
 
@@
     system, designed for supporting knowledge engineers in the process
-    of writing, analyzing and expanding PDDL files and thereby
+    of writing, analyzing and expanding \pddl files and thereby
     promoting the collaboration between knowledge engineers and the
-    use of PDDL in real-world applications. It consists of the parts
+    use of \pddl in real-world applications. It consists of the parts
     shown in the figure.}
@@
 mismatch between domain and problem file in real time. The code
-completion feature offers recommendations for standard standard
-\textsc{pddl} constructs as well as for previously used terms. Code
-folding allows the knowledge engineer to hide currently not needed
-code blocks. In this case only the first line of the block is
-displayed. Lastly, a command-line interface allows the integration of
-planners in order to run and compare different planning software.
+completion feature offers recommendations for standard \textsc{pddl}
+constructs as well as for previously used terms. Code folding allows
+the knowledge engineer to hide currently not needed code blocks. In
+this case only the first line of the block is displayed. Lastly, a
+command-line interface allows the integration of planners in order to
+run and compare different planning software.
 
@@
 The online tool \epd allows for editing \pddl files in a web
-browser. Its feature comprise syntax highlighting, code folding,
+browser. Its features comprise syntax highlighting, code folding,
 \pddl-specific auto-completion, and multi-tab support. The editor is
@@
 custom key shortcuts, customizing the general look and feel, and
-bracket matching. In contrast, \vscode is integrated into the
-mature editor Visual Studio Code and can be used in combination with
-other plugins. To have both basic editor features and a high
+bracket matching. In contrast, \vscode and \pddlmode for Emacs are
+integrated into mature code editors and can be used in combination
+with other plugins. To have both basic editor features and a high
 customizability, it was decided to use an existing, extensible text
@@
 
-Another feature that can be useful for fast development is the ability
-to insert larger code skeletons or snippets. \pddlstudio does not
-support the insertion of code snippets. \itsimple features some code
-templates for predicates, derived predicates, functions, actions,
-constraints, types, comments, requirements, objects, and
-metrics. However, the templates are neither customizable nor
-extensible. \pddlmode for Emacs provides three larger skeletons: one
-for domains, one for problems, and one for actions. Further skeletons
-could be added. Both \epd and \vscode provide many code
-snippets. \mypddl aims to combine the best of these latter tools and
-support customizable and extensible snippets for domains, problems,
-types, predicates, functions, actions, and durative actions.
+Another useful feature for fast development is the ability to insert
+larger code skeletons or snippets. \pddlstudio does not support the
+insertion of code snippets. \itsimple features some code templates for
+predicates, derived predicates, functions, actions, constraints,
+types, comments, requirements, objects, and metrics. However, the
+templates are neither customizable nor extensible. \pddlmode for Emacs
+provides three larger skeletons: one for domains, one for problems,
+and one for actions. Further skeletons could be added. Both \epd and
+\vscode provide many code snippets. \mypddl aims to combine the best
+of these latter tools and support customizable and extensible snippets
+for domains, problems, types, predicates, functions, actions, and
+durative actions.
 
@@
 extensible. This is the other major reason why it was decided that
-\mypddl should be integrated into a existing, extensible, and
+\mypddl should be integrated into an existing, extensible, and
 customizable text editor. These requirements are met by Sublime Text,
@@
 trigger (\texttt{p2} would insert the binary predicate template
-\texttt{(pred-name~?x~-~object~?y~-~object)}.
+\texttt{(pred-name~?x~-~object~?y~-~object)}).
 
@@
 \mypddldiagram, since they support collaboration, efficiency, and
-debugging independently of the user’s experience with \pddl. To
+debugging independently of the user's experience with \pddl. To
 evaluate their usability, they were evaluated in a user study. To this
@@
 24 hours before the experiment was to take place, participants
-received the web
-link\footnote{Tutorial in German: \url{https://www.youtube.com/watch?v=Uck-K8VnNOU&list=PL3CZzLUZuiIMWEfJxy-G6OxYVzUrvjwuV}} to a 30-minute interactive video tutorial
-on AI planning and \pddl. This method was chosen in order not to
-pressure the participant with the presence of an experimenter when
-trying to understand the material.
+received the web link\footnote{Tutorial in German:
+  \url{https://www.youtube.com/watch?v=Uck-K8VnNOU&list=PL3CZzLUZuiIMWEfJxy-G6OxYVzUrvjwuV}}
+to a 30-minute interactive video tutorial on AI planning and
+\pddl. This method was chosen in order not to pressure the participant
+with the presence of an experimenter when trying to understand the
+material.
 
@@
 
-We defined four tasks: two debugging tasks for testing the syntax
-highlighting feature and two type hierarchy tasks for testing the type
-diagram generator. A within subjects design was considered most suited
-due to the small number of participants. Therefore, it was necessary to
-construct two tasks matched in difficulty for each of these two
-types to compare the effects of having the tools available. Each
-participant started either with a debugging or type hierarchy task and
-was given the \mypddl tools either in the first two tasks or the
-second two tasks, so that each participant completed each task type
-once with and once without \mypddl. This results in $2$ (first task is
-debugging or hierarchy) $\times$ $2$ (task variations for debugging
-and hierarchy) $\times$ $2$ (starting with or without \mypddl) $= 8$
-individual task orders, one per participant.
+We defined four tasks (Appendix~\ref{sub:tasks}): two debugging tasks
+for testing the syntax highlighting feature and two type hierarchy
+tasks for testing the type diagram generator. A within subjects design
+was considered most suited due to the small number of
+participants. Therefore, it was necessary to construct two tasks
+matched in difficulty for each of these two types to compare the
+effects of having the tools available. Each participant started either
+with a debugging or type hierarchy task and was given the \mypddl
+tools either in the first two tasks or the second two tasks, so that
+each participant completed each task type once with and once without
+\mypddl. This results in $2$ (first task is debugging or hierarchy)
+$\times$ $2$ (task variations for debugging and hierarchy) $\times$
+$2$ (starting with or without \mypddl) $= 8$ individual task orders,
+one per participant.
 
@@
 
+\newpage
 \subsubsection{Analysis}
-
 \begin{itemize}
-\item Debugging Tasks: \nopagebreak To compare differences in the
+\item Debugging Tasks: To compare differences in the
   debugging tasks, a paired sample $t$-test was used; normality was
@@
   The participants detected more errors using the syntax highlighting
-  feature, ($M = 7.6$, $SD = 2.07$) than without it ($M = 10.3$,
-  $SD = 3.45$); $t(7) = 2.68$, $p = 0.03$.  That is, approximately
+  feature ($M = 10.3$, $SD = 3.45$) than without it ($M = 7.6$,
+  $SD = 2.07$); $t(7) = 2.68$, $p = 0.03$.  That is, approximately
   36\,\% more errors were found with syntax highlighting. The
@@
     \end{tabular}
-
     \caption{Overview of geometric means ($GM$s), geometric standard deviations ($GSD$s), degrees of freedom ($df$), $t$-values, and $p$-values.
@@
 In contrast, Figure \ref{fig:task-completions-agg} displays the geometric means for all correct answers.}
-
     \label{tab:p-values}
@@
 \begin{figure}[h]
-\captionsetup{singlelinecheck=off}
   \centering
@@
   found all errors in the domain, while none achieved this without
-  syntax highlighting.
+  syntax highlighting. While every participant had to use the same
+  color scheme in the experiment, colors are customizable in
+  \sublimetext.
 
